---
title: "Sequence Analysis of App Usage Data"
subtitle: "Method 1 — TraMineR · ICA 2026 Hackathon"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    self-contained: true
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(TraMineR)
library(TraMineRextras)
library(colorspace)
library(cluster)
library(patchwork)
```

## Data

**Source:** `uni_mainz_app_session.csv` (~20M rows).

Relevant columns (full codebook in `Readme.txt`):

| Column             | Type    | Notes                                              |
|--------------------|---------|----------------------------------------------------|
| `participant_code` | char    | person identifier; use for linkage with survey     |
| `app_name`         | char    | displayed name (e.g., "Kalender", "Instagram")     |
| `package_name`     | char    | unique Android package (e.g., `com.instagram.android`) |
| `start_time`       | POSIXct | app session start — use this, **not** `seen_timestamp` |
| `end_time`         | POSIXct | app session end — use this, **not** `end_timestamp` |
| `Duration`         | numeric | seconds of app use (pre-computed, 3 decimal places)|
| `phone_session_id` | numeric | increments each phone unlock; resets at some range |
| `app_session_id`   | numeric | app-level flow ID; **has gaps** for blacklisted apps (e.g., WhatsApp) |

> **Timestamp note:** `seen_timestamp` and `end_timestamp` are epoch/UNIX with
> a ~2 h offset (CEST summer time). Always use `start_time` / `end_time`.

> **Scale note:** 20M rows makes full-sample TraMineR infeasible (OM distance
> is O(n²) in persons). The pipeline below samples to a manageable subset for
> clustering; run descriptives on the full data first.

### App categorisation

`app_category` is not in the raw data — we derive it from `package_name`.
Extend the lookup table below once the actual package names are known.

### Load data

```{r load-data}
# --- Real data (uncomment when available) ---
# app_category comes from a project-provided lookup file (package_name → app_category)
# app_categories <- read_csv("data/raw/app_categories.csv")  # provided by project
#
# df_raw <- read_csv(
#   "data/raw/uni_mainz_app_session.csv",
#   col_types = cols(
#     participant_code = col_character(),
#     start_time       = col_datetime(),
#     end_time         = col_datetime(),
#     Duration         = col_double(),
#     phone_session_id = col_double(),
#     app_session_id   = col_double(),
#     package_name     = col_character(),
#     app_name         = col_character(),
#     .default         = col_skip()      # skip id, user_id, seen_timestamp, etc.
#   )
# ) |>
#   left_join(app_categories, by = "package_name") |>
#   arrange(participant_code, start_time)

# --- SYNTHETIC SAMPLE for development ---
set.seed(42)

pkg_to_app <- c(
  "com.instagram.android"         = "Instagram",
  "com.whatsapp"                  = "WhatsApp",
  "com.zhiliaoapp.musically"      = "TikTok",
  "com.google.android.youtube"    = "YouTube",
  "com.android.chrome"            = "Chrome",
  "com.google.android.apps.maps"  = "Maps",
  "com.google.android.calculator" = "Rechner",
  "com.google.android.gm"         = "Gmail",
  "com.facebook.orca"             = "Messenger",
  "com.spotify.music"             = "Spotify"
)
pkg_to_cat <- c(
  "com.instagram.android"         = "social network",
  "com.whatsapp"                  = "communication",
  "com.zhiliaoapp.musically"      = "social network",
  "com.google.android.youtube"    = "social network",
  "com.android.chrome"            = "tools",
  "com.google.android.apps.maps"  = "tools",
  "com.google.android.calculator" = "tools",
  "com.google.android.gm"         = "communication",
  "com.facebook.orca"             = "communication",
  "com.spotify.music"             = "entertainment"
)
pkg_names <- names(pkg_to_app)
pkg_probs  <- c(3, 3, 2, 2, 1, 1, .5, 1, 1, 1)

n_persons <- 50
events_per <- sample(80:400, n_persons, replace = TRUE)

df_raw <- map2_dfr(
  paste0("P", sprintf("%03d", seq_len(n_persons))),
  events_per,
  function(pid, n) {
    starts <- as.POSIXct("2024-04-01 08:00:00") +
      cumsum(c(0, sample(5:1800, n - 1, replace = TRUE)))
    pkgs <- sample(pkg_names, n, replace = TRUE, prob = pkg_probs)
    durs <- sample(5:900, n, replace = TRUE) + runif(n)
    tibble(
      participant_code = pid,
      package_name     = pkgs,
      app_name         = pkg_to_app[pkgs],
      app_category     = pkg_to_cat[pkgs],
      start_time       = starts,
      end_time         = starts + durs,
      Duration         = durs,
      phone_session_id = cumsum(c(0, as.numeric(diff(starts)) > 300)) %% 100,
      app_session_id   = ifelse(pkgs == "com.whatsapp", NA_real_, seq_along(pkgs))
    )
  }
) |> arrange(participant_code, start_time)
# --- end synthetic sample ---
```

---

## Data preparation

### Step 1 — Sanity checks

```{r sanity-checks}
# Duration should match end_time - start_time within rounding
df_raw |>
  slice_sample(n = 1000) |>
  mutate(
    derived_dur = as.numeric(end_time - start_time, units = "secs"),
    dur_diff    = abs(Duration - derived_dur)
  ) |>
  summarise(max_discrepancy_sec = max(dur_diff, na.rm = TRUE),
            mean_discrepancy_sec = mean(dur_diff, na.rm = TRUE)) |>
  knitr::kable(caption = "Duration column vs. derived from timestamps (sample of 1000)")

# app_session_id gaps (blacklisted apps)
n_missing_session <- sum(is.na(df_raw$app_session_id))
cat(sprintf(
  "Missing app_session_id: %d rows (%.1f%%) — likely blacklisted apps (e.g. WhatsApp)\n",
  n_missing_session, 100 * n_missing_session / nrow(df_raw)
))

# phone_session_id range
cat(sprintf(
  "phone_session_id range: %d – %d\n",
  min(df_raw$phone_session_id, na.rm = TRUE),
  max(df_raw$phone_session_id, na.rm = TRUE)
))
```

### Step 2 — Filter and clean

```{r clean}
df <- df_raw |>
  filter(
    !is.na(participant_code),
    !is.na(start_time),
    Duration > 0      # drop zero-duration artefacts
  ) |>
  # Use Duration directly; convert to seconds if needed elsewhere
  mutate(duration_sec = Duration) |>
  arrange(participant_code, start_time)
```

### Step 3 — Merge consecutive same-app episodes

Consecutive episodes on the same app, separated by < `gap_threshold` seconds,
are collapsed. This follows the 10-minute rule in Fan et al. (2026).
Because `Duration` is pre-computed, we sum durations rather than re-derive.

```{r merge-episodes}
gap_threshold <- 10 * 60  # seconds — DECISION POINT: inspect gaps in real data

merge_consecutive <- function(d, gap_sec) {
  d <- arrange(d, start_time)
  out  <- list()
  cur  <- d[1, ]
  for (i in seq_len(nrow(d))[-1]) {
    row <- d[i, ]
    gap <- as.numeric(row$start_time - cur$end_time, units = "secs")
    if (row$app_name == cur$app_name && !is.na(gap) && gap <= gap_sec) {
      cur$end_time     <- row$end_time
      cur$duration_sec <- cur$duration_sec + row$duration_sec
    } else {
      out <- c(out, list(cur))
      cur <- row
    }
  }
  bind_rows(c(out, list(cur)))
}

df_merged <- df |>
  group_by(participant_code) |>
  group_modify(~ merge_consecutive(.x, gap_threshold)) |>
  ungroup()
```

### Step 4 — Sequence length distribution

Inspect before deciding how many persons / how long sequences to use for TraMineR.

```{r seq-lengths}
seq_len_dist <- df_merged |>
  count(participant_code, name = "seq_length")

seq_len_dist |>
  summarise(
    n_persons = n(),
    min = min(seq_length), q25 = quantile(seq_length, .25),
    median = median(seq_length), q75 = quantile(seq_length, .75),
    max = max(seq_length)
  ) |>
  knitr::kable(caption = "Sequence lengths (turns per person, after merging)")

# CCDF plot
seq_len_sorted <- sort(seq_len_dist$seq_length)
tibble(
  length = seq_len_sorted,
  ccdf   = 1 - seq_along(seq_len_sorted) / length(seq_len_sorted)
) |>
  ggplot(aes(length, ccdf)) +
  geom_line() +
  scale_x_log10() + scale_y_log10() +
  labs(x = "Sequence length (turns)", y = "CCDF",
       title = "CCDF of per-person sequence lengths") +
  theme_minimal()
```

### Step 5 — Filter to IQR and sample for TraMineR

TraMineR's OM distance is O(n²) in persons and O(L²) in sequence length.
For ~20M rows, work on a sample; run descriptive analyses on the full data.

```{r filter-sample}
# Keep persons with sequence lengths in IQR (mirrors reference study)
q25 <- quantile(seq_len_dist$seq_length, .25)
q75 <- quantile(seq_len_dist$seq_length, .75)

eligible <- seq_len_dist |>
  filter(seq_length >= q25, seq_length <= q75) |>
  pull(participant_code)

# DECISION POINT: TraMineR is feasible up to ~300-500 persons.
# Sample down if more eligible persons exist.
max_persons_traminer <- 300
sampled_persons <- if (length(eligible) > max_persons_traminer) {
  set.seed(1)
  sample(eligible, max_persons_traminer)
} else {
  eligible
}

df_filtered <- df_merged |>
  filter(participant_code %in% sampled_persons) |>
  group_by(participant_code) |>
  mutate(turn = row_number()) |>
  ungroup()

cat(sprintf(
  "Using %d persons, sequence lengths %d–%d turns\n",
  length(sampled_persons), q25, q75
))
```

### Step 6 — Reshape to wide format

```{r wide-format}
max_turn <- max(df_filtered$turn)

# App-level wide matrix
df_wide_app <- df_filtered |>
  select(participant_code, turn, app_name) |>
  pivot_wider(names_from = turn, values_from = app_name, names_prefix = "t") |>
  select(participant_code, paste0("t", seq_len(max_turn)))

# Category-level wide matrix
df_wide_cat <- df_filtered |>
  select(participant_code, turn, app_category) |>
  pivot_wider(names_from = turn, values_from = app_category, names_prefix = "t") |>
  select(participant_code, paste0("t", seq_len(max_turn)))
```

---

## Sequence analysis — app level

### Alphabet and colour palette

```{r alphabet-app}
app_alphabet <- sort(unique(df_filtered$app_name))
n_apps <- length(app_alphabet)
app_pal <- qualitative_hcl(n_apps, palette = "Dark 3")
names(app_pal) <- app_alphabet

# DECISION POINT: replace with brand colours once app list is known.
# Known overrides:
overrides <- c(
  "Instagram" = "#E1306C", "WhatsApp"  = "#25D366",
  "TikTok"    = "#010101", "YouTube"   = "#FF0000",
  "Chrome"    = "#4285F4", "Gmail"     = "#EA4335",
  "Spotify"   = "#1DB954", "Messenger" = "#0084FF"
)
for (nm in intersect(names(overrides), names(app_pal))) {
  app_pal[nm] <- overrides[nm]
}

cat("App alphabet:\n"); print(app_alphabet)
```

### Create sequence object

```{r seqdef-app}
app.seq <- seqdef(
  df_wide_app |> select(-participant_code),
  alphabet = app_alphabet,
  labels   = app_alphabet,
  cpal     = app_pal[app_alphabet],
  xtstep   = max(1, round(max_turn / 20))
)
```

### Visualisations

```{r seqplots-app, fig.width=12, fig.height=8}
seqIplot(app.seq, sortv = "from.start", with.legend = "right",
         cex.legend = 0.6, main = "App usage sequences (sampled persons)")
```

```{r seqfplot-app, fig.width=10, fig.height=5}
seqfplot(app.seq, with.legend = "right", border = NA,
         main = "Most frequent distinct sequences")
```

```{r seqdplot-app, fig.width=10, fig.height=5}
seqdplot(app.seq, with.legend = "right", border = NA,
         main = "State distribution over sequence position")
```

```{r seqtrate-app}
seqtrate(app.seq) |> round(3) |>
  knitr::kable(caption = "Transition rates between apps")
```

---

## Sequence analysis — category level

```{r seqdef-cat}
cat_alphabet <- sort(unique(df_filtered$app_category))
cat_pal <- c(
  "social network" = "#F28E2B",
  "communication"  = "#4E79A7",
  "tools"          = "#59A14F",
  "entertainment"  = "#B07AA1",
  "other"          = "#BAB0AC"
)
missing_cats <- setdiff(cat_alphabet, names(cat_pal))
if (length(missing_cats)) cat_pal[missing_cats] <- gray.colors(length(missing_cats))

cat.seq <- seqdef(
  df_wide_cat |> select(-participant_code),
  alphabet = cat_alphabet,
  labels   = cat_alphabet,
  cpal     = cat_pal[cat_alphabet]
)
```

```{r seqplots-cat, fig.width=12, fig.height=6}
seqIplot(cat.seq, sortv = "from.start", with.legend = "right",
         main = "App-category sequences")
seqdplot(cat.seq, with.legend = "right", border = NA,
         main = "Category distribution over sequence position")
```

---

## N-gram motif analysis

Run on the **full merged data** (not just the TraMineR sample) to maximise
statistical power. Tests bigrams and trigrams against an independence baseline
with Bonferroni correction (α = 0.001), following Fan et al.

```{r ngrams}
# Use full eligible set, not just TraMineR sample
sequences_full <- df_merged |>
  filter(participant_code %in% eligible) |>
  arrange(participant_code, start_time) |>
  group_by(participant_code) |>
  summarise(seq = list(app_name), .groups = "drop")

get_ngrams <- function(seq_list, n) {
  seq_list |>
    lapply(\(s) if (length(s) >= n) {
      lapply(seq_len(length(s) - n + 1), \(i) s[i:(i + n - 1)])
    } else NULL) |>
    unlist(recursive = FALSE) |>
    sapply(paste, collapse = " -> ")
}

bigrams  <- get_ngrams(sequences_full$seq, 2)
trigrams <- get_ngrams(sequences_full$seq, 3)

base_freq <- prop.table(table(df_merged$app_name[
  df_merged$participant_code %in% eligible
]))

test_ngrams <- function(ngram_vec, base_freq, alpha = 0.001) {
  counts     <- sort(table(ngram_vec), decreasing = TRUE)
  total      <- sum(counts)
  bonf_alpha <- alpha / length(counts)
  bind_rows(lapply(names(counts), function(ng) {
    states <- strsplit(ng, " -> ")[[1]]
    obs_p  <- counts[ng] / total
    exp_p  <- prod(base_freq[states], na.rm = FALSE)
    if (is.na(exp_p) || exp_p == 0) return(NULL)
    sd_p   <- sqrt(exp_p * (1 - exp_p) / total)
    z      <- (obs_p - exp_p) / sd_p
    p      <- 2 * pnorm(-abs(z))
    if (p < bonf_alpha)
      tibble(ngram = ng, observed = counts[ng],
             obs_pct = obs_p * 100, exp_pct = exp_p * 100,
             z_score = z, p_value = p)
  })) |> arrange(desc(z_score))
}

bigram_sig  <- test_ngrams(bigrams,  base_freq)
trigram_sig <- test_ngrams(trigrams, base_freq)

knitr::kable(head(bigram_sig,  10), digits = 3,
             caption = "Top over-represented bigrams (Bonferroni p < 0.001)")
```

```{r trigrams-table}
knitr::kable(head(trigram_sig, 10), digits = 3,
             caption = "Top over-represented trigrams (Bonferroni p < 0.001)")
```

---

## Optimal matching distance & clustering

```{r om-distance}
costmatrix <- seqsubm(app.seq, method = "CONSTANT", cval = 2, with.missing = TRUE)

# SCALE NOTE: slow for > 300 persons. Cache the result.
dist_om_path <- "results/tables/dist_om_app.rds"
if (file.exists(dist_om_path)) {
  dist_om <- readRDS(dist_om_path)
} else {
  dist_om <- seqdist(app.seq, method = "OM", indel = 1,
                     sm = costmatrix, with.missing = TRUE)
  saveRDS(dist_om, dist_om_path)
}
```

```{r clustering}
clust_ward <- agnes(dist_om, diss = TRUE, method = "ward")
plot(clust_ward, which.plot = 2, main = "Dendrogram — app sequences")
```

```{r silhouette, fig.width=7, fig.height=4}
k_range    <- 2:8
sil_widths <- sapply(k_range, \(k) mean(silhouette(cutree(clust_ward, k), dist_om)[, 3]))

tibble(k = k_range, avg_silhouette = sil_widths) |>
  ggplot(aes(k, avg_silhouette)) +
  geom_line() + geom_point() +
  labs(title = "Average silhouette width by k",
       x = "Number of clusters", y = "Avg silhouette") +
  theme_minimal()
```

```{r cluster-viz, fig.width=12, fig.height=8}
k_chosen <- 3  # DECISION POINT: update after inspecting silhouette above
clusters  <- cutree(clust_ward, k = k_chosen)

seqIplot(app.seq, group = clusters, sortv = "from.start",
         with.legend = "right", cex.legend = 0.6,
         main = paste0("Clusters (k = ", k_chosen, ")"))
```

```{r cluster-sizes}
tibble(participant_code = sampled_persons, cluster = as.character(clusters)) |>
  count(cluster) |>
  knitr::kable(caption = "Persons per cluster")
```

---

## Phone-session-level analysis

`phone_session_id` increments each time the phone is unlocked. Analysing
*within-session* sequences answers a different question: given a single
phone pickup, in what order are apps used?

```{r session-analysis}
# Sequence = apps used within one phone session (one unlock episode)
# Unit of analysis: session (not person)
df_sessions <- df_merged |>
  # Use full data here, not just TraMineR sample
  arrange(participant_code, start_time) |>
  group_by(participant_code, phone_session_id) |>
  mutate(turn_in_session = row_number()) |>
  ungroup()

session_lengths <- df_sessions |>
  count(participant_code, phone_session_id, name = "session_length")

session_lengths |>
  summarise(
    n_sessions = n(),
    median_length = median(session_length),
    q75 = quantile(session_length, .75),
    max_length = max(session_length)
  ) |>
  knitr::kable(caption = "Phone session lengths (apps per unlock)")

# NOTE on phone_session_id: the codebook notes it is a two-digit numeric value
# with an unclear range/reset logic. Verify that IDs don't wrap within a person
# before treating them as globally unique session boundaries.
```
