---
title: "Sequence Analysis of App Usage Data"
subtitle: "Method 1 — TraMineR · ICA 2026 Hackathon"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    self-contained: true
execute:
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---

```{r setup}
library(tidyverse)
library(TraMineR)
library(TraMineRextras)
library(colorspace)
library(cluster)
library(patchwork)
library(here)
source(here::here("R", "utils.R"))
```

## Data

**Source:** `uni_mainz_app_session.csv` (~20M rows).

Relevant columns (full codebook in `Readme.txt`):

| Column             | Type    | Notes                                              |
|--------------------|---------|----------------------------------------------------|
| `participant_code` | char    | person identifier; use for linkage with survey     |
| `app_name`         | char    | displayed name (e.g., "Kalender", "Instagram")     |
| `package_name`     | char    | unique Android package (e.g., `com.instagram.android`) |
| `start_time`       | POSIXct | app session start — use this, **not** `seen_timestamp` |
| `end_time`         | POSIXct | app session end — use this, **not** `end_timestamp` |
| `Duration`         | numeric | seconds of app use (pre-computed, 3 decimal places)|
| `phone_session_id` | numeric | increments each phone unlock; resets at some range |
| `app_session_id`   | numeric | app-level flow ID; **has gaps** for blacklisted apps (e.g., WhatsApp) |

> **Timestamp note:** `seen_timestamp` and `end_timestamp` are epoch/UNIX with
> a ~2 h offset (CEST summer time). Always use `start_time` / `end_time`.

> **Unit of analysis:** one sequence per **person-day** — all merged app episodes
> for one participant on one calendar day. This keeps sequences short (bounded
> by waking hours) and positions time-comparable across persons. OM distance
> is then between days, not persons; person-level patterns emerge from the
> cluster distribution across each participant's days.
>
> **Scale note:** with the full dataset (~1000 persons × 46 days ≈ 46 000
> sequences) OM is still O(n²) in sequences. Sample to ≤ 500 sequences for
> clustering; run descriptives on the full set first.

### App categorisation

`app_category` is not in the raw data — we derive it from `package_name`.
Extend the lookup table below once the actual package names are known.

### Load data

```{r load-data}
df_raw <- read_csv(
  here::here("_files/d_apps_anonymized.csv"),
  col_types = cols(
    participant_code = col_character(),
    start_time       = col_datetime(),
    end_time         = col_datetime(),
    duration         = col_double(),
    phone_session_id = col_double(),
    app_session_id   = col_double(),
    package_name     = col_character(),
    app_name         = col_character(),
    category         = col_character(),
    .default         = col_skip()
  )
) |>
  rename(app_category = category, Duration = duration) |>
  mutate(app_category = translate_category(app_category)) |>
  apply_category_overrides() |>
  arrange(participant_code, start_time)
```

---

## Data preparation

### Step 1 — Sanity checks

```{r sanity-checks}
# Duration should match end_time - start_time within rounding
df_raw |>
  slice_sample(n = 1000) |>
  mutate(
    derived_dur = as.numeric(end_time - start_time, units = "secs"),
    dur_diff    = abs(Duration - derived_dur)
  ) |>
  summarise(max_discrepancy_sec = max(dur_diff, na.rm = TRUE),
            mean_discrepancy_sec = mean(dur_diff, na.rm = TRUE)) |>
  knitr::kable(caption = "Duration column vs. derived from timestamps (sample of 1000)")

# app_session_id gaps (blacklisted apps)
n_missing_session <- sum(is.na(df_raw$app_session_id))
cat(sprintf(
  "Missing app_session_id: %f rows (%.1f%%) — likely blacklisted apps (e.g. WhatsApp)\n",
  n_missing_session, 100 * n_missing_session / nrow(df_raw)
))

# phone_session_id range
cat(sprintf(
  "phone_session_id range: %f – %f\n",
  min(df_raw$phone_session_id, na.rm = TRUE),
  max(df_raw$phone_session_id, na.rm = TRUE)
))
```

### Step 2 — Filter and clean

```{r clean}
df <- df_raw |>
  filter(
    !is.na(participant_code),
    !is.na(start_time),
    Duration > 0      # drop zero-duration artefacts
  ) |>
  arrange(participant_code, start_time)
```

### Step 3 — Merge consecutive same-app episodes

Consecutive episodes on the same app, separated by < `gap_threshold` seconds,
are collapsed. This follows the 10-minute rule in Fan et al. (2026).
Because `Duration` is pre-computed, we sum durations rather than re-derive.

```{r merge-episodes}
df_merged <- df |>
  group_by(participant_code) |>
  group_modify(~ merge_consecutive(.x, gap_threshold)) |>
  ungroup() |>
  filter(!app_name %in% noise_apps)  # remove launcher/home-screen artefacts (see utils.R)
```

### Step 4 — Person-day sequence length distribution

Each person-day is one sequence. Inspect lengths to choose `min_len` and `max_len`.

```{r seq-lengths}
seq_len_dist <- df_merged |>
  mutate(date = as.Date(start_time)) |>
  count(participant_code, date, name = "seq_length")

seq_len_dist |>
  summarise(
    n_person_days = n(),
    n_persons     = n_distinct(participant_code),
    min    = min(seq_length),
    q25    = quantile(seq_length, .25),
    median = median(seq_length),
    q75    = quantile(seq_length, .75),
    max    = max(seq_length)
  ) |>
  knitr::kable(caption = "Events per person-day (after merging)")

seq_len_dist |>
  ggplot(aes(seq_length)) +
  geom_histogram(bins = 30, fill = "#4E79A7", colour = "white") +
  labs(x = "Events per day", y = "Person-days",
       title = "Distribution of daily sequence lengths (all participants)") +
  theme_minimal()
```

### Step 5 — Filter and sample person-days for TraMineR

Drop very short days (too little information) and cap very long ones.
At scale, sample up to `max_seq_traminer` person-days for OM clustering.

```{r filter-sample}
min_len <- 5   # minimum merged events per day
max_len <- 50  # cap: sequences truncated to this length

valid_days <- seq_len_dist |> filter(seq_length >= min_len)

max_seq_traminer <- 500  # OM is O(n²); keep ≤ 500 for feasibility
set.seed(1)
sampled_days <- if (nrow(valid_days) > max_seq_traminer) {
  slice_sample(valid_days, n = max_seq_traminer)
} else {
  valid_days
}

df_filtered <- df_merged |>
  mutate(date = as.Date(start_time)) |>
  inner_join(sampled_days |> select(participant_code, date),
             by = c("participant_code", "date")) |>
  group_by(participant_code, date) |>
  mutate(turn = row_number()) |>
  filter(turn <= max_len) |>
  ungroup()

cat(sprintf(
  "%d person-days retained across %d persons; sequences capped at %d turns\n",
  nrow(sampled_days), n_distinct(sampled_days$participant_code), max_len
))

# Reduce alphabet: collapse apps with fewer than min_app_obs episodes across all
# person-days into "(other app)". This keeps the legend and seqIplot readable.
min_app_obs <- 10  # DECISION POINT — lower = more detail, higher = cleaner plots
app_episode_counts <- table(df_filtered$app_name)
frequent_apps <- names(app_episode_counts)[app_episode_counts >= min_app_obs]

df_filtered <- df_filtered |>
  mutate(app_name = if_else(app_name %in% frequent_apps, app_name, "(other app)"))

cat(sprintf(
  "Alphabet: %d frequent apps kept; %d rare apps collapsed into '(other app)'\n",
  length(frequent_apps), sum(app_episode_counts < min_app_obs)
))
```

### Step 6 — Reshape to wide format

One row per person-day. Columns t1 … t`max_len` are the app sequence for that day.

```{r wide-format}
max_turn <- max_len  # already filtered in step 5

# App-level wide matrix (rows = person-days)
df_wide_app <- df_filtered |>
  select(participant_code, date, turn, app_name) |>
  pivot_wider(names_from = turn, values_from = app_name, names_prefix = "t") |>
  select(participant_code, date, starts_with("t"))

# Category-level wide matrix
df_wide_cat <- df_filtered |>
  select(participant_code, date, turn, app_category) |>
  pivot_wider(names_from = turn, values_from = app_category, names_prefix = "t") |>
  select(participant_code, date, starts_with("t"))

cat(sprintf("TraMineR input: %d person-days, sequences up to %d turns\n",
            nrow(df_wide_app), max_turn))
```

---

## Sequence analysis — app level

### Alphabet and colour palette

```{r alphabet-app}
app_alphabet <- sort(unique(df_filtered$app_name[!is.na(df_filtered$app_name)]))
n_apps <- length(app_alphabet)
app_pal <- qualitative_hcl(n_apps, palette = "Dark 3")
names(app_pal) <- app_alphabet

# DECISION POINT: replace with brand colours once app list is known.
# Known overrides:
overrides <- c(
  "Instagram" = "#E1306C", "WhatsApp"  = "#25D366",
  "TikTok"    = "#010101", "YouTube"   = "#FF0000",
  "Chrome"    = "#4285F4", "Gmail"     = "#EA4335",
  "Spotify"   = "#1DB954", "Messenger" = "#0084FF"
)
for (nm in intersect(names(overrides), names(app_pal))) {
  app_pal[nm] <- overrides[nm]
}

cat("App alphabet:\n"); print(app_alphabet)
```

### Create sequence object

```{r seqdef-app}
app.seq <- seqdef(
  df_wide_app |> select(-participant_code, -date),
  alphabet = app_alphabet,
  labels   = app_alphabet,
  cpal     = app_pal[app_alphabet],
  xtstep   = max(1, round(max_turn / 20))
)
```

### Visualisations

```{r seqplots-app, fig.width=12, fig.height=8}
tryCatch(
  seqIplot(app.seq, sortv = NULL, with.legend = FALSE,
           cex.legend = 0.6, main = "App usage sequences (person-days)"),
  error = function(e) message("seqIplot: ", conditionMessage(e))
)
```

```{r seqfplot-app, fig.width=10, fig.height=5}
tryCatch(
  seqfplot(app.seq, with.legend = FALSE, border = NA,
           main = "Most frequent distinct sequences"),
  error = function(e) message("seqfplot: ", conditionMessage(e))
)
```

State distribution at the app level is unreadable with 50+ states; see the
category-level `seqdplot` in the section below for the interpretable version.

```{r seqtrate-cat-inline}
# App-level seqtrate would produce a 116×116 table — show category-level instead.
# (Category-level seqtrate is repeated here; the full cat.seq section below has plots.)
tryCatch({
  cat_alphabet_tmp <- sort(unique(df_filtered$app_category))
  df_wide_cat_tmp  <- df_filtered |>
    select(participant_code, date, turn, app_category) |>
    pivot_wider(names_from = turn, values_from = app_category, names_prefix = "t") |>
    select(participant_code, date, starts_with("t"))
  cat.seq.tmp <- seqdef(
    df_wide_cat_tmp |> select(-participant_code, -date),
    alphabet = cat_alphabet_tmp, labels = cat_alphabet_tmp
  )
  seqtrate(cat.seq.tmp) |> round(3) |>
    knitr::kable(caption = "Transition rates between app categories")
}, error = function(e) message("seqtrate: ", conditionMessage(e)))
```

---

## Sequence analysis — category level

```{r seqdef-cat}
cat_alphabet <- sort(unique(df_filtered$app_category))
cat_pal <- c(
  "communication"  = "#4E79A7",
  "social media"   = "#F28E2B",
  "entertainment"  = "#B07AA1",
  "gaming"         = "#E15759",
  "utilities"      = "#59A14F",
  "shopping"       = "#76B7B2",
  "practical life" = "#EDC948"
)
missing_cats <- setdiff(cat_alphabet, names(cat_pal))
if (length(missing_cats)) cat_pal[missing_cats] <- gray.colors(length(missing_cats))

cat.seq <- seqdef(
  df_wide_cat |> select(-participant_code, -date),
  alphabet = cat_alphabet,
  labels   = cat_alphabet,
  cpal     = cat_pal[cat_alphabet]
)
```

```{r seqplots-cat, fig.width=12, fig.height=6}
tryCatch(
  seqIplot(cat.seq, sortv = NULL, with.legend = FALSE,
           main = "App-category sequences"),
  error = function(e) message("seqIplot cat: ", conditionMessage(e))
)
tryCatch(
  seqdplot(cat.seq, with.legend = FALSE, border = NA,
           main = "Category distribution over sequence position"),
  error = function(e) message("seqdplot cat: ", conditionMessage(e))
)
```

---

## N-gram motif analysis

Run on **all merged data** (not just the TraMineR sample) to maximise
statistical power. Tests bigrams and trigrams against an independence baseline
with Bonferroni correction (α = 0.001), following Fan et al.

```{r ngrams}
sequences_full <- df_merged |>
  arrange(participant_code, start_time) |>
  group_by(participant_code) |>
  summarise(seq = list(app_name), .groups = "drop")

get_ngrams <- function(seq_list, n) {
  seq_list |>
    lapply(\(s) if (length(s) >= n) {
      lapply(seq_len(length(s) - n + 1), \(i) s[i:(i + n - 1)])
    } else NULL) |>
    unlist(recursive = FALSE) |>
    sapply(paste, collapse = " -> ")
}

bigrams  <- get_ngrams(sequences_full$seq, 2)
trigrams <- get_ngrams(sequences_full$seq, 3)

# Drop self-loops (A→A, A→A→B, etc.) — merge_consecutive handles short-gap
# returns; remaining same-app bigrams are trivial and dominate the ranking.
has_self_loop <- function(ngram_vec) {
  vapply(strsplit(ngram_vec, " -> "), \(s) any(s[-length(s)] == s[-1]), logical(1))
}
bigrams  <- bigrams[!has_self_loop(bigrams)]
trigrams <- trigrams[!has_self_loop(trigrams)]

base_freq <- prop.table(table(df_merged$app_name))

# Independence null: P(A)×P(B) for bigrams, P(A)×P(B)×P(C) for trigrams.
# Tests whether consecutive co-occurrence exceeds chance given marginal frequencies.
# A Markov null (P(A)×P(B|A)) is circular when estimated from the same data —
# the conditional probability already encodes the observed transition, forcing exp≈obs.
test_ngrams <- function(ngram_vec, base_freq, alpha = 0.001, max_ngrams = 5000) {
  counts     <- sort(table(ngram_vec), decreasing = TRUE)
  total      <- sum(counts)
  bonf_alpha <- alpha / length(counts)
  counts     <- head(counts, max_ngrams)
  result <- bind_rows(lapply(names(counts), function(ng) {
    states <- strsplit(ng, " -> ")[[1]]
    obs_p  <- counts[ng] / total
    exp_p  <- prod(base_freq[states])
    if (is.na(exp_p) || exp_p == 0) return(NULL)
    sd_p <- sqrt(exp_p * (1 - exp_p) / total)
    z    <- (obs_p - exp_p) / sd_p
    p    <- 2 * pnorm(-abs(z))
    if (p < bonf_alpha)
      tibble(ngram = ng, observed = counts[ng],
             obs_pct = obs_p * 100, exp_pct = exp_p * 100,
             z_score = z, p_value = p)
  }))
  if (nrow(result) == 0 || !"z_score" %in% names(result))
    return(tibble(ngram = character(), observed = integer(),
                  obs_pct = numeric(), exp_pct = numeric(),
                  z_score = numeric(), p_value = numeric()))
  result |> arrange(desc(z_score))
}

bigram_sig  <- test_ngrams(bigrams,  base_freq)
trigram_sig <- test_ngrams(trigrams, base_freq)

knitr::kable(head(bigram_sig,  10), digits = 3,
             caption = "Top over-represented bigrams (Bonferroni p < 0.001)")
```

```{r trigrams-table}
knitr::kable(head(trigram_sig, 10), digits = 3,
             caption = "Top over-represented trigrams (Bonferroni p < 0.001)")
```

---

## Optimal matching distance & clustering

```{r om-distance}
costmatrix <- seqsubm(app.seq, method = "TRATE", with.missing = TRUE)

# SCALE NOTE: slow for > 500 sequences. Cache the result.
dist_om_path <- "results/tables/dist_om_personday.rds"
if (file.exists(dist_om_path)) {
  dist_om <- readRDS(dist_om_path)
} else {
  dist_om <- seqdist(app.seq, method = "OM", indel = 1,
                     sm = costmatrix, with.missing = TRUE)
  saveRDS(dist_om, dist_om_path)
}
```

```{r clustering}
n_seq <- nrow(dist_om)
if (n_seq >= 3) {
  clust_ward <- agnes(dist_om, diss = TRUE, method = "ward")
} else {
  message("Fewer than 3 sequences — skipping clustering (needs full dataset)")
  clust_ward <- NULL
}
```

```{r silhouette, fig.width=7, fig.height=4}
if (!is.null(clust_ward)) {
  k_range    <- 2:min(8, n_seq - 1)
  sil_widths <- sapply(k_range, \(k) mean(silhouette(cutree(clust_ward, k), dist_om)[, 3]))
  tibble(k = k_range, avg_silhouette = sil_widths) |>
    ggplot(aes(k, avg_silhouette)) +
    geom_line() + geom_point() +
    labs(title = "Average silhouette width by k",
         x = "Number of clusters", y = "Avg silhouette") +
    theme_minimal()
}
```

```{r cluster-viz, fig.width=12, fig.height=8}
if (!is.null(clust_ward)) {
  k_chosen <- k_range[which.max(sil_widths)]
  clusters  <- cutree(clust_ward, k = k_chosen)
  tryCatch(
    seqIplot(app.seq, group = clusters, sortv = NULL,
             with.legend = FALSE, cex.legend = 0.6,
             main = paste0("Clusters (k = ", k_chosen, ")")),
    error = function(e) message("seqIplot clusters: ", conditionMessage(e))
  )
} else {
  clusters <- NULL
}
```

```{r cluster-sizes}
if (!is.null(clusters)) {
  day_clusters <- df_wide_app |>
    select(participant_code, date) |>
    mutate(cluster = as.character(clusters))

  day_clusters |>
    count(cluster) |>
    knitr::kable(caption = "Person-days per cluster")

  # How are each person's days distributed across cluster types?
  day_clusters |>
    count(participant_code, cluster) |>
    group_by(participant_code) |>
    mutate(pct = round(100 * n / sum(n), 1)) |>
    ungroup() |>
    select(participant_code, cluster, days = n, pct) |>
    arrange(participant_code, cluster) |>
    knitr::kable(caption = "Cluster distribution per person (% of their days)")
}
```

---

## Optimal matching distance & clustering — category level

Category-level sequences (5 states) produce more readable sequence plots and
more interpretable cluster labels than app-level sequences. This section
repeats the OM → clustering pipeline on `cat.seq`.

```{r om-distance-cat}
dist_om_cat_path <- "results/tables/dist_om_cat.rds"
if (file.exists(dist_om_cat_path)) {
  dist_om_cat <- readRDS(dist_om_cat_path)
} else {
  costmatrix_cat <- seqsubm(cat.seq, method = "TRATE", with.missing = TRUE)
  dist_om_cat    <- seqdist(cat.seq, method = "OM", indel = 1,
                            sm = costmatrix_cat, with.missing = TRUE)
  saveRDS(dist_om_cat, dist_om_cat_path)
}
```

```{r clustering-cat}
n_seq_cat <- nrow(dist_om_cat)
if (n_seq_cat >= 3) {
  clust_ward_cat <- agnes(dist_om_cat, diss = TRUE, method = "ward")
} else {
  message("Fewer than 3 sequences — skipping category clustering")
  clust_ward_cat <- NULL
}
```

```{r silhouette-cat, fig.width=7, fig.height=4}
if (!is.null(clust_ward_cat)) {
  k_range_cat    <- 2:min(8, n_seq_cat - 1)
  sil_widths_cat <- sapply(k_range_cat,
    \(k) mean(silhouette(cutree(clust_ward_cat, k), dist_om_cat)[, 3]))
  tibble(k = k_range_cat, avg_silhouette = sil_widths_cat) |>
    ggplot(aes(k, avg_silhouette)) +
    geom_line() + geom_point() +
    labs(title = "Average silhouette width by k (category sequences)",
         x = "Number of clusters", y = "Avg silhouette") +
    theme_minimal()
}
```

```{r cluster-viz-cat, fig.width=12, fig.height=8}
if (!is.null(clust_ward_cat)) {
  k_chosen_cat  <- k_range_cat[which.max(sil_widths_cat)]
  clusters_cat  <- cutree(clust_ward_cat, k = k_chosen_cat)
  tryCatch(
    seqIplot(cat.seq, group = clusters_cat, sortv = NULL,
             with.legend = TRUE, cex.legend = 0.8,
             main = paste0("Category clusters (k = ", k_chosen_cat, ")")),
    error = function(e) message("seqIplot cat clusters: ", conditionMessage(e))
  )
}
```

---

## Phone-session-level analysis (exploratory)

Complementary to the person-day sequences above. A phone session = one unlock
episode; within-session sequences ask: given a single pickup, in what order
are apps used? This is a finer-grained unit than person-day.

**Caveat:** `phone_session_id` is not well-documented (README marks it "xxx").
The IDs show non-monotonic patterns in the data and appear to reset irregularly.
Use for grouping only — do not treat the ID values as ordinal or as a reliable
timeline. Verify with the data provider before publishing session-level results.

```{r session-analysis}
# Sequence = apps used within one phone session (one unlock episode)
# Unit of analysis: session (not person)
df_sessions <- df_merged |>
  # Use full data here, not just TraMineR sample
  arrange(participant_code, start_time) |>
  group_by(participant_code, phone_session_id) |>
  mutate(turn_in_session = row_number()) |>
  ungroup()

session_lengths <- df_sessions |>
  count(participant_code, phone_session_id, name = "session_length")

session_lengths |>
  summarise(
    n_sessions = n(),
    median_length = median(session_length),
    q75 = quantile(session_length, .75),
    max_length = max(session_length)
  ) |>
  knitr::kable(caption = "Phone session lengths (apps per unlock)")

# NOTE on phone_session_id: the codebook notes it is a two-digit numeric value
# with an unclear range/reset logic. Verify that IDs don't wrap within a person
# before treating them as globally unique session boundaries.
```
