---
title: "Event History Analysis of App Usage"
subtitle: "Method 2 — Survival Analysis · ICA 2026 Hackathon"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    self-contained: true
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(survival)
library(survminer)
library(broom)
library(scales)
library(splines)
library(car)
source("utils.R")
```

## Conceptual framing

The core question: **how long does a person use an app before switching to
a different one?**

Each app-use episode (after merging same-app consecutive uses) is one row:

- **Time** (`T`): episode duration in minutes (`Duration / 60`)
- **Event** (`E`): did the person immediately switch to a *different* app?
  - `E = 1` → app switch observed
  - `E = 0` → censored (last episode per person, or last in session)

The reference study (Fan et al., 2026) applied this to platform switching
across Facebook/YouTube/TikTok/Instagram. We apply it at the **app level**
and the coarser **app-category level**.

Our data has two structural advantages the reference lacked:

1. **`Duration` is pre-computed** — no need to derive from events.
2. **`phone_session_id` is meaningful** — each value = one phone unlock.
   We can model whether hazard of switching differs *within* vs. *across*
   sessions, or by position within a session.

---

## Data

**Source:** `uni_mainz_app_session.csv` (~20M rows).

```{r load-data}
df_raw <- read_csv(
  "../_files/d_apps_anonymized.csv",
  col_types = cols(
    participant_code = col_character(),
    start_time       = col_datetime(),
    end_time         = col_datetime(),
    duration         = col_double(),
    phone_session_id = col_double(),
    app_session_id   = col_double(),
    package_name     = col_character(),
    app_name         = col_character(),
    category         = col_character(),
    .default         = col_skip()
  )
) |>
  rename(app_category = category, Duration = duration) |>
  mutate(app_category = translate_category(app_category)) |>
  arrange(participant_code, start_time)
```

---

## Data preparation

### Step 1 — Merge consecutive same-app episodes

Same 10-minute rule as Method 1. When merging, durations are summed.

```{r merge-episodes}
df_merged <- df_raw |>
  filter(!is.na(participant_code), Duration > 0) |>
  group_by(participant_code) |>
  group_modify(~ merge_consecutive(.x, gap_threshold)) |>
  ungroup() |>
  filter(Duration >= 2) |>           # drop sub-2-second launcher/system artefacts
  filter(!app_name %in% noise_apps)  # remove launcher/home-screen artefacts (see utils.R)
```

### Step 2 — Define switch events

```{r define-events}
df_events <- df_merged |>
  mutate(duration_min = Duration / 60) |>
  arrange(participant_code, start_time) |>
  group_by(participant_code) |>
  mutate(
    next_app        = lead(app_name),
    next_session_id = lead(phone_session_id),
    next_category   = lead(app_category),
    # Observed switch: different app AND within the same phone session.
    # Episodes ending a session are censored (phone put down = not an app switch).
    app_switch = as.integer(
      !is.na(next_app) &
      next_app != app_name &
      next_session_id == phone_session_id
    ),
    cat_switch = as.integer(
      !is.na(next_category) &
      next_category != app_category &
      next_session_id == phone_session_id
    ),
    app_switch = replace_na(app_switch, 0),
    cat_switch = replace_na(cat_switch, 0),
    episode_in_session = row_number(),  # within person; refine per session below
  ) |>
  ungroup()

# NOTE: session-boundary censoring relies on phone_session_id reliability.
# The codebook notes non-monotonic IDs and unclear reset logic — verify with
# the data provider before publishing session-boundary-dependent results.
cat(sprintf(
  "Rows with missing app_session_id (blacklisted apps): %d (%.1f%%)\n",
  sum(is.na(df_events$app_session_id)),
  100 * mean(is.na(df_events$app_session_id))
))
```

### Step 3 — Descriptive overview

```{r desc-summary}
df_events |>
  summarise(
    n_persons    = n_distinct(participant_code),
    n_episodes   = n(),
    n_app_switch = sum(app_switch),
    pct_switch   = mean(app_switch) * 100,
    med_dur_min  = median(duration_min),
    mean_dur_min = mean(duration_min)
  ) |>
  knitr::kable(digits = 1, caption = "Episode-level summary")
```

---

## Session duration distributions

CCDF of episode durations by app — mirrors Figure 1 of Fan et al.
Run on **all** episodes (no sampling needed here).

```{r ccdf-by-app, fig.width=9, fig.height=5}
min_obs <- 50  # minimum episodes to draw a CCDF curve
apps_to_plot <- df_events |> count(app_name) |> filter(n >= min_obs) |> pull(app_name)

df_events |>
  filter(app_name %in% apps_to_plot) |>
  group_by(app_name) |>
  arrange(duration_min, .by_group = TRUE) |>
  mutate(ccdf = 1 - row_number() / n()) |>
  ungroup() |>
  ggplot(aes(duration_min, ccdf, colour = app_name)) +
  geom_line(linewidth = 0.7) +
  scale_x_log10(labels = label_number(suffix = " min")) +
  scale_y_log10() +
  labs(x = "Duration (minutes, log)", y = "CCDF",
       title = "CCDF of app session durations", colour = "App") +
  theme_minimal()
```

```{r ccdf-by-category, fig.width=8, fig.height=4}
cat_pal <- c(
  "social network" = "#F28E2B", "communication" = "#4E79A7",
  "tools" = "#59A14F", "entertainment" = "#B07AA1", "other" = "#BAB0AC"
)

df_events |>
  filter(app_category %in% names(cat_pal)) |>
  group_by(app_category) |>
  arrange(duration_min, .by_group = TRUE) |>
  mutate(ccdf = 1 - row_number() / n()) |>
  ungroup() |>
  ggplot(aes(duration_min, ccdf, colour = app_category)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = cat_pal) +
  scale_x_log10(labels = label_number(suffix = " min")) +
  scale_y_log10() +
  labs(x = "Duration (minutes, log)", y = "CCDF",
       title = "CCDF by app category", colour = "Category") +
  theme_minimal()
```

---

## Kaplan-Meier survival analysis

### Overall: time until any app switch

```{r km-overall}
km_overall <- survfit(Surv(duration_min, app_switch) ~ 1, data = df_events)

ggsurvplot(
  km_overall, data = df_events,
  conf.int = TRUE,
  xlab = "Duration (minutes)",
  ylab = "P(staying on same app)",
  title = "Overall: time until app switch",
  ggtheme = theme_minimal()
)

surv_tab <- summary(km_overall)$table
cat(sprintf(
  "Median time until app switch: %.1f min (95%% CI: %.1f–%.1f)\n",
  surv_tab["median"], surv_tab["0.95LCL"], surv_tab["0.95UCL"]
))
```

### Stratified by app

```{r km-by-app, fig.height=7}
km_app <- survfit(
  Surv(duration_min, app_switch) ~ app_name,
  data = filter(df_events, app_name %in% apps_to_plot)
)

ggsurvplot(
  km_app,
  data         = filter(df_events, app_name %in% apps_to_plot),
  conf.int     = FALSE,
  pval         = TRUE, pval.method = TRUE,
  xlab         = "Duration (minutes)",
  ylab         = "P(staying on same app)",
  title        = "Time until app switch, by app",
  legend.title = "App",
  xlim         = c(0, quantile(df_events$duration_min, 0.95)),
  ggtheme      = theme_minimal()
)
```

```{r km-medians}
tidy(km_app) |>
  filter(estimate <= 0.5) |>
  group_by(strata) |>
  slice_min(time, n = 1) |>
  ungroup() |>
  transmute(app = str_remove(strata, "app_name="), median_min = time) |>
  arrange(median_min) |>
  knitr::kable(caption = "Median time (min) on each app before switching")
```

### Stratified by app category

```{r km-by-category}
km_cat <- survfit(
  Surv(duration_min, cat_switch) ~ app_category,
  data = df_events
)

tryCatch(
  print(ggsurvplot(
    km_cat, data = df_events,
    conf.int = TRUE, pval = TRUE, pval.method = TRUE,
    xlab = "Duration (minutes)",
    ylab = "P(staying in same category)",
    title = "Time until category switch, by category",
    legend.title = "Category",
    ggtheme = theme_minimal()
  )),
  error = function(e) {
    message("ggsurvplot conf.int incompatible with this ggplot2 version — rendering without CI")
    print(ggsurvplot(
      km_cat, data = df_events,
      conf.int = FALSE, pval = TRUE, pval.method = TRUE,
      xlab = "Duration (minutes)",
      ylab = "P(staying in same category)",
      title = "Time until category switch, by category",
      legend.title = "Category",
      ggtheme = theme_minimal()
    ))
  }
)
```

### Log-rank tests

```{r logrank}
survdiff(Surv(duration_min, app_switch) ~ app_name,
         data = filter(df_events, app_name %in% apps_to_plot)) |> print()

survdiff(Surv(duration_min, cat_switch) ~ app_category,
         data = df_events) |> print()
```

---

## Cox Proportional Hazards model

### Feature engineering

```{r cox-features}
# Compute episode position within each phone session first, then join on
episode_pos <- df_events |>
  group_by(participant_code, phone_session_id) |>
  mutate(episode_in_session = row_number()) |>
  ungroup() |>
  select(participant_code, phone_session_id, start_time,
         episode_in_session_sess = episode_in_session)

df_cox <- df_events |>
  left_join(episode_pos,
            by = c("participant_code", "phone_session_id", "start_time")) |>
  mutate(
    episode_in_session = episode_in_session_sess,
    hour_of_day  = as.integer(format(start_time, "%H")),
    app_name     = factor(app_name),
    app_category = factor(app_category),
    day_of_week  = factor(weekdays(as.Date(start_time)))
  )

# M2/M3 restrict to top 20 apps — fitting 100+ app dummies on 860K rows
# does not converge in reasonable time and the tail apps carry little signal.
top20_apps <- df_cox |> count(app_name, sort = TRUE) |> slice_head(n = 20) |> pull(app_name)
df_cox_top <- df_cox |> filter(app_name %in% top20_apps) |> mutate(app_name = droplevels(app_name))
```

### Model 1 — Category only

```{r cox-m1}
m1 <- coxph(Surv(duration_min, app_switch) ~ app_category, data = df_cox)

tidy(m1, exponentiate = TRUE, conf.int = TRUE) |>
  select(term, HR = estimate, CI_low = conf.low, CI_high = conf.high, p.value) |>
  knitr::kable(digits = 3, caption = "Cox M1: hazard ratios by app category")
```

### Model 2 — App + time of day + weekday

```{r cox-m2}
m2 <- coxph(
  Surv(duration_min, app_switch) ~ app_name + ns(hour_of_day, df = 4) + day_of_week,
  data = df_cox_top
)

tidy(m2, exponentiate = TRUE, conf.int = TRUE) |>
  select(term, HR = estimate, CI_low = conf.low, CI_high = conf.high, p.value) |>
  knitr::kable(digits = 3, caption = "Cox M2: app + time covariates")
```

### Multicollinearity check (VIF)

Before interpreting M2 coefficients, confirm that app, time, and weekday
covariates do not carry redundant information (e.g., certain apps used only
on weekends, or only at specific hours).

```{r vif-m2}
# car::vif() returns GVIF and GVIF^(1/(2*Df)) for multi-df terms (factors, splines).
# Rule of thumb: GVIF^(1/(2*Df)) > 2 indicates problematic collinearity.
vif(m2) |>
  as.data.frame() |>
  rownames_to_column("term") |>
  knitr::kable(digits = 2, caption = "Generalised VIF — Cox M2 covariates")
```

```{r cox-forest, fig.height=5}
# Forest plot on M1 (category level): M2 has ~115 app terms and is unreadable as a forest plot.
tryCatch(
  ggforest(m1, data = df_cox, main = "Hazard ratios for app switching — by category (Cox M1)"),
  error = function(e) message("ggforest: ", conditionMessage(e))
)
```

### Model 3 — Add phone session position

Session position tests whether the hazard of switching rises (or falls) as
the person uses more apps within a single phone-unlock episode.

```{r cox-m3}
m3 <- coxph(
  Surv(duration_min, app_switch) ~ app_name + ns(hour_of_day, df = 4) + day_of_week +
    episode_in_session,
  data = df_cox_top
)

# Likelihood-ratio comparison
anova(m2, m3) |> knitr::kable(caption = "LRT: does session position improve fit?")

tidy(m3, exponentiate = TRUE, conf.int = TRUE) |>
  filter(term == "episode_in_session") |>
  select(term, HR = estimate, CI_low = conf.low, CI_high = conf.high, p.value) |>
  knitr::kable(digits = 3, caption = "Session position effect on switching hazard")
```

### Proportional hazards assumption

```{r ph-test}
# M1 (category only): clean diagnostic — 5 terms, interpretable plots
ph_test_m1 <- cox.zph(m1)
print(ph_test_m1)
ggcoxzph(ph_test_m1)

# M2 (app + time + weekday): 115+ terms — print table only, skip ggcoxzph
ph_test_m2 <- cox.zph(m2)
print(ph_test_m2)
```

---

## Phone session analysis

`phone_session_id` = phone unlock episode. This lets us ask: how does
switching behaviour differ by session characteristics?

```{r session-km, fig.width=8, fig.height=4}
df_session <- df_events |>
  group_by(participant_code, phone_session_id) |>
  mutate(
    session_turn   = row_number(),
    is_first_in_session = session_turn == 1
  ) |>
  ungroup()

# KM: first app of a session vs. subsequent apps within same session
km_session <- survfit(
  Surv(duration_min, app_switch) ~ is_first_in_session,
  data = df_session
)

tryCatch(
  print(ggsurvplot(
    km_session, data = df_session,
    conf.int = TRUE, pval = TRUE,
    legend.labs = c("Later in session", "First app of session"),
    xlab = "Duration (minutes)",
    title = "First vs. later app within a phone session",
    ggtheme = theme_minimal()
  )),
  error = function(e) {
    message("ggsurvplot conf.int incompatible with this ggplot2 version — rendering without CI")
    print(ggsurvplot(
      km_session, data = df_session,
      conf.int = FALSE, pval = TRUE,
      legend.labs = c("Later in session", "First app of session"),
      xlab = "Duration (minutes)",
      title = "First vs. later app within a phone session",
      ggtheme = theme_minimal()
    ))
  }
)
```

```{r session-length-effect}
# Does session length (how many apps total) predict per-episode switch rate?
session_stats <- df_session |>
  group_by(participant_code, phone_session_id) |>
  summarise(
    session_length = n(),
    switch_rate    = mean(app_switch),
    .groups = "drop"
  )

session_stats |>
  ggplot(aes(session_length, switch_rate)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_smooth(method = "loess", se = TRUE) +
  scale_x_log10() +
  labs(
    x = "Session length (apps used, log scale)",
    y = "Within-session app switch rate",
    title = "Does session length predict switching frequency?"
  ) +
  theme_minimal()
```

---

## Summary of key outputs

| Output              | What it shows                                           |
|---------------------|---------------------------------------------------------|
| CCDF by app         | How heavy-tailed session durations are per app          |
| CCDF by category    | Duration distributions: social vs. communication vs. tools |
| KM overall          | Median time until any app switch                        |
| KM by app           | Which apps hold attention longest                       |
| KM by category      | Category-level switching speed                          |
| Log-rank test       | Whether group survival curves differ significantly      |
| VIF (M2)            | Whether app, time, and weekday covariates are collinear |
| Cox M1              | Category-level hazard ratios                            |
| Cox M2              | App-level HRs + time of day + weekday                  |
| Cox M3              | Whether episode position within a session matters       |
| PH assumption test  | Whether Cox model assumptions hold                      |
| Session KM          | First app of pickup vs. later apps                      |
| Session length plot | Whether longer pickups have higher switch rates         |
