---
title: "Event History Analysis of App Usage"
subtitle: "Method 2 — Survival Analysis · ICA 2026 Hackathon"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    self-contained: true
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(survival)
library(survminer)
library(broom)
library(scales)
```

## Conceptual framing

The core question: **how long does a person use an app before switching to
a different one?**

Each app-use episode (after merging same-app consecutive uses) is one row:

- **Time** (`T`): episode duration in minutes (`Duration / 60`)
- **Event** (`E`): did the person immediately switch to a *different* app?
  - `E = 1` → app switch observed
  - `E = 0` → censored (last episode per person, or last in session)

The reference study (Fan et al., 2026) applied this to platform switching
across Facebook/YouTube/TikTok/Instagram. We apply it at the **app level**
and the coarser **app-category level**.

Our data has two structural advantages the reference lacked:

1. **`Duration` is pre-computed** — no need to derive from events.
2. **`phone_session_id` is meaningful** — each value = one phone unlock.
   We can model whether hazard of switching differs *within* vs. *across*
   sessions, or by position within a session.

---

## Data

**Source:** `uni_mainz_app_session.csv` (~20M rows).

```{r load-data}
# --- Real data (uncomment when available) ---
# app_category comes from a project-provided lookup file (package_name → app_category)
# app_categories <- read_csv("data/raw/app_categories.csv")  # provided by project
#
# df_raw <- read_csv(
#   "data/raw/uni_mainz_app_session.csv",
#   col_types = cols(
#     participant_code = col_character(),
#     start_time       = col_datetime(),
#     end_time         = col_datetime(),
#     Duration         = col_double(),
#     phone_session_id = col_double(),
#     app_session_id   = col_double(),
#     package_name     = col_character(),
#     app_name         = col_character(),
#     .default         = col_skip()
#   )
# ) |>
#   left_join(app_categories, by = "package_name") |>
#   arrange(participant_code, start_time)

# --- SYNTHETIC SAMPLE for development ---
set.seed(42)

pkg_to_app <- c(
  "com.instagram.android"         = "Instagram",
  "com.whatsapp"                  = "WhatsApp",
  "com.zhiliaoapp.musically"      = "TikTok",
  "com.google.android.youtube"    = "YouTube",
  "com.android.chrome"            = "Chrome",
  "com.google.android.apps.maps"  = "Maps",
  "com.google.android.calculator" = "Rechner",
  "com.google.android.gm"         = "Gmail",
  "com.facebook.orca"             = "Messenger",
  "com.spotify.music"             = "Spotify"
)
pkg_to_cat <- c(
  "com.instagram.android"         = "social network",
  "com.whatsapp"                  = "communication",
  "com.zhiliaoapp.musically"      = "social network",
  "com.google.android.youtube"    = "social network",
  "com.android.chrome"            = "tools",
  "com.google.android.apps.maps"  = "tools",
  "com.google.android.calculator" = "tools",
  "com.google.android.gm"         = "communication",
  "com.facebook.orca"             = "communication",
  "com.spotify.music"             = "entertainment"
)
pkg_names <- names(pkg_to_app)
pkg_probs  <- c(3, 3, 2, 2, 1, 1, .5, 1, 1, 1)

n_persons <- 80
events_per <- sample(50:200, n_persons, replace = TRUE)

df_raw <- map2_dfr(
  paste0("P", sprintf("%03d", seq_len(n_persons))),
  events_per,
  function(pid, n) {
    starts <- as.POSIXct("2024-04-01 08:00:00") +
      cumsum(c(0, sample(5:1800, n - 1, replace = TRUE)))
    pkgs <- sample(pkg_names, n, replace = TRUE, prob = pkg_probs)
    durs <- sample(5:900, n, replace = TRUE) + runif(n)
    tibble(
      participant_code = pid,
      package_name     = pkgs,
      app_name         = pkg_to_app[pkgs],
      app_category     = pkg_to_cat[pkgs],
      start_time       = starts,
      end_time         = starts + durs,
      Duration         = durs,
      phone_session_id = cumsum(c(0, as.numeric(diff(starts)) > 300)) %% 100,
      app_session_id   = ifelse(pkgs == "com.whatsapp", NA_real_, seq_along(pkgs))
    )
  }
) |> arrange(participant_code, start_time)
# --- end synthetic sample ---
```

---

## Data preparation

### Step 1 — Merge consecutive same-app episodes

Same 10-minute rule as Method 1. When merging, durations are summed.

```{r merge-episodes}
gap_threshold <- 10 * 60  # seconds

merge_consecutive <- function(d, gap_sec) {
  d <- arrange(d, start_time)
  out <- list()
  cur <- d[1, ]
  for (i in seq_len(nrow(d))[-1]) {
    row <- d[i, ]
    gap <- as.numeric(row$start_time - cur$end_time, units = "secs")
    if (row$app_name == cur$app_name && !is.na(gap) && gap <= gap_sec) {
      cur$end_time <- row$end_time
      cur$Duration <- cur$Duration + row$Duration
    } else {
      out <- c(out, list(cur))
      cur <- row
    }
  }
  bind_rows(c(out, list(cur)))
}

df_merged <- df_raw |>
  filter(!is.na(participant_code), Duration > 0) |>
  group_by(participant_code) |>
  group_modify(~ merge_consecutive(.x, gap_threshold)) |>
  ungroup()
```

### Step 2 — Define switch events

```{r define-events}
df_events <- df_merged |>
  mutate(duration_min = Duration / 60) |>
  group_by(participant_code) |>
  mutate(
    app_switch = as.integer(!is.na(lead(app_name)) & lead(app_name) != app_name),
    cat_switch = as.integer(!is.na(lead(app_category)) & lead(app_category) != app_category),
    # Last episode per person: no observed switch → right-censored
    app_switch = replace_na(app_switch, 0),
    cat_switch = replace_na(cat_switch, 0),
    # Position in phone session (meaningful: phone_session_id = unlock event)
    episode_in_session = row_number(),  # within person; refine per session below
  ) |>
  ungroup()

# NOTE on app_session_id: gaps exist where apps are blacklisted (e.g., WhatsApp).
# This does NOT affect the survival analysis, which is based on app_name not app_session_id.
# But it means we may undercount true app switches where WhatsApp is involved.
cat(sprintf(
  "Rows with missing app_session_id (blacklisted apps): %d (%.1f%%)\n",
  sum(is.na(df_events$app_session_id)),
  100 * mean(is.na(df_events$app_session_id))
))
```

### Step 3 — Descriptive overview

```{r desc-summary}
df_events |>
  summarise(
    n_persons    = n_distinct(participant_code),
    n_episodes   = n(),
    n_app_switch = sum(app_switch),
    pct_switch   = mean(app_switch) * 100,
    med_dur_min  = median(duration_min),
    mean_dur_min = mean(duration_min)
  ) |>
  knitr::kable(digits = 1, caption = "Episode-level summary")
```

---

## Session duration distributions

CCDF of episode durations by app — mirrors Figure 1 of Fan et al.
Run on **all** episodes (no sampling needed here).

```{r ccdf-by-app, fig.width=9, fig.height=5}
min_obs <- 50  # minimum episodes to draw a CCDF curve
apps_to_plot <- df_events |> count(app_name) |> filter(n >= min_obs) |> pull(app_name)

df_events |>
  filter(app_name %in% apps_to_plot) |>
  group_by(app_name) |>
  arrange(duration_min, .by_group = TRUE) |>
  mutate(ccdf = 1 - row_number() / n()) |>
  ungroup() |>
  ggplot(aes(duration_min, ccdf, colour = app_name)) +
  geom_line(linewidth = 0.7) +
  scale_x_log10(labels = label_number(suffix = " min")) +
  scale_y_log10() +
  labs(x = "Duration (minutes, log)", y = "CCDF",
       title = "CCDF of app session durations", colour = "App") +
  theme_minimal()
```

```{r ccdf-by-category, fig.width=8, fig.height=4}
cat_pal <- c(
  "social network" = "#F28E2B", "communication" = "#4E79A7",
  "tools" = "#59A14F", "entertainment" = "#B07AA1", "other" = "#BAB0AC"
)

df_events |>
  filter(app_category %in% names(cat_pal)) |>
  group_by(app_category) |>
  arrange(duration_min, .by_group = TRUE) |>
  mutate(ccdf = 1 - row_number() / n()) |>
  ungroup() |>
  ggplot(aes(duration_min, ccdf, colour = app_category)) +
  geom_line(linewidth = 0.9) +
  scale_colour_manual(values = cat_pal) +
  scale_x_log10(labels = label_number(suffix = " min")) +
  scale_y_log10() +
  labs(x = "Duration (minutes, log)", y = "CCDF",
       title = "CCDF by app category", colour = "Category") +
  theme_minimal()
```

---

## Kaplan-Meier survival analysis

### Overall: time until any app switch

```{r km-overall}
km_overall <- survfit(Surv(duration_min, app_switch) ~ 1, data = df_events)

ggsurvplot(
  km_overall, data = df_events,
  conf.int = TRUE,
  xlab = "Duration (minutes)",
  ylab = "P(staying on same app)",
  title = "Overall: time until app switch",
  ggtheme = theme_minimal()
)

surv_tab <- summary(km_overall)$table
cat(sprintf(
  "Median time until app switch: %.1f min (95%% CI: %.1f–%.1f)\n",
  surv_tab["median"], surv_tab["0.95LCL"], surv_tab["0.95UCL"]
))
```

### Stratified by app

```{r km-by-app, fig.height=7}
km_app <- survfit(
  Surv(duration_min, app_switch) ~ app_name,
  data = filter(df_events, app_name %in% apps_to_plot)
)

ggsurvplot(
  km_app,
  data         = filter(df_events, app_name %in% apps_to_plot),
  conf.int     = FALSE,
  pval         = TRUE, pval.method = TRUE,
  xlab         = "Duration (minutes)",
  ylab         = "P(staying on same app)",
  title        = "Time until app switch, by app",
  legend.title = "App",
  xlim         = c(0, quantile(df_events$duration_min, 0.95)),
  ggtheme      = theme_minimal()
)
```

```{r km-medians}
tidy(km_app) |>
  filter(estimate <= 0.5) |>
  group_by(strata) |>
  slice_min(time, n = 1) |>
  ungroup() |>
  transmute(app = str_remove(strata, "app_name="), median_min = time) |>
  arrange(median_min) |>
  knitr::kable(caption = "Median time (min) on each app before switching")
```

### Stratified by app category

```{r km-by-category}
km_cat <- survfit(
  Surv(duration_min, cat_switch) ~ app_category,
  data = df_events
)

ggsurvplot(
  km_cat, data = df_events,
  conf.int = TRUE, pval = TRUE, pval.method = TRUE,
  palette  = unname(cat_pal[sort(unique(df_events$app_category))]),
  xlab = "Duration (minutes)",
  ylab = "P(staying in same category)",
  title = "Time until category switch, by category",
  legend.title = "Category",
  ggtheme = theme_minimal()
)
```

### Log-rank tests

```{r logrank}
survdiff(Surv(duration_min, app_switch) ~ app_name,
         data = filter(df_events, app_name %in% apps_to_plot)) |> print()

survdiff(Surv(duration_min, cat_switch) ~ app_category,
         data = df_events) |> print()
```

---

## Cox Proportional Hazards model

### Feature engineering

```{r cox-features}
# Compute episode position within each phone session first, then join on
episode_pos <- df_events |>
  group_by(participant_code, phone_session_id) |>
  mutate(episode_in_session = row_number()) |>
  ungroup() |>
  select(participant_code, phone_session_id, start_time, episode_in_session)

df_cox <- df_events |>
  left_join(episode_pos,
            by = c("participant_code", "phone_session_id", "start_time")) |>
  mutate(
    hour_of_day  = as.integer(format(start_time, "%H")),
    day_of_week  = weekdays(as.Date(start_time)),
    app_name     = relevel(factor(app_name),     ref = "Chrome"),
    app_category = relevel(factor(app_category), ref = "tools"),
    day_of_week  = relevel(factor(day_of_week),  ref = "Monday")
  )
```

### Model 1 — Category only

```{r cox-m1}
m1 <- coxph(Surv(duration_min, app_switch) ~ app_category, data = df_cox)

tidy(m1, exponentiate = TRUE, conf.int = TRUE) |>
  select(term, HR = estimate, CI_low = conf.low, CI_high = conf.high, p.value) |>
  knitr::kable(digits = 3, caption = "Cox M1: hazard ratios by app category")
```

### Model 2 — App + time of day + weekday

```{r cox-m2}
# DECISION POINT: add/remove predictors to match your RQ.
# hour_of_day: linear here; use splines (splines::ns()) if effect is non-linear.
m2 <- coxph(
  Surv(duration_min, app_switch) ~ app_name + hour_of_day + day_of_week,
  data = df_cox
)

tidy(m2, exponentiate = TRUE, conf.int = TRUE) |>
  select(term, HR = estimate, CI_low = conf.low, CI_high = conf.high, p.value) |>
  knitr::kable(digits = 3, caption = "Cox M2: app + time covariates")
```

```{r cox-forest, fig.height=7}
ggforest(m2, data = df_cox, main = "Hazard ratios for app switching (Cox M2)")
```

### Model 3 — Add phone session position

Session position tests whether the hazard of switching rises (or falls) as
the person uses more apps within a single phone-unlock episode.

```{r cox-m3}
m3 <- coxph(
  Surv(duration_min, app_switch) ~ app_name + hour_of_day + day_of_week +
    episode_in_session,
  data = df_cox
)

# Likelihood-ratio comparison
anova(m2, m3) |> knitr::kable(caption = "LRT: does session position improve fit?")

tidy(m3, exponentiate = TRUE, conf.int = TRUE) |>
  filter(term == "episode_in_session") |>
  select(term, HR = estimate, CI_low = conf.low, CI_high = conf.high, p.value) |>
  knitr::kable(digits = 3, caption = "Session position effect on switching hazard")
```

### Proportional hazards assumption

```{r ph-test}
ph_test <- cox.zph(m2)
print(ph_test)
ggcoxzph(ph_test)
```

---

## Phone session analysis

`phone_session_id` = phone unlock episode. This lets us ask: how does
switching behaviour differ by session characteristics?

```{r session-km, fig.width=8, fig.height=4}
df_session <- df_events |>
  group_by(participant_code, phone_session_id) |>
  mutate(
    session_turn   = row_number(),
    is_first_in_session = session_turn == 1
  ) |>
  ungroup()

# KM: first app of a session vs. subsequent apps within same session
km_session <- survfit(
  Surv(duration_min, app_switch) ~ is_first_in_session,
  data = df_session
)

ggsurvplot(
  km_session, data = df_session,
  conf.int = TRUE, pval = TRUE,
  legend.labs = c("Later in session", "First app of session"),
  xlab = "Duration (minutes)",
  title = "First vs. later app within a phone session",
  ggtheme = theme_minimal()
)
```

```{r session-length-effect}
# Does session length (how many apps total) predict per-episode switch rate?
session_stats <- df_session |>
  group_by(participant_code, phone_session_id) |>
  summarise(
    session_length = n(),
    switch_rate    = mean(app_switch),
    .groups = "drop"
  )

session_stats |>
  ggplot(aes(session_length, switch_rate)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_smooth(method = "loess", se = TRUE) +
  scale_x_log10() +
  labs(
    x = "Session length (apps used, log scale)",
    y = "Within-session app switch rate",
    title = "Does session length predict switching frequency?"
  ) +
  theme_minimal()
```

---

## Summary of key outputs

| Output              | What it shows                                           |
|---------------------|---------------------------------------------------------|
| CCDF by app         | How heavy-tailed session durations are per app          |
| CCDF by category    | Duration distributions: social vs. communication vs. tools |
| KM overall          | Median time until any app switch                        |
| KM by app           | Which apps hold attention longest                       |
| KM by category      | Category-level switching speed                          |
| Log-rank test       | Whether group survival curves differ significantly      |
| Cox M1              | Category-level hazard ratios                            |
| Cox M2              | App-level HRs + time of day + weekday                  |
| Cox M3              | Whether episode position within a session matters       |
| PH assumption test  | Whether Cox model assumptions hold                      |
| Session KM          | First app of pickup vs. later apps                      |
| Session length plot | Whether longer pickups have higher switch rates         |
