---
title: "Hidden Markov Models of App Usage"
subtitle: "Method 3 — depmixS4 · ICA 2026 Hackathon"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    self-contained: true
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(depmixS4)
library(scales)
library(patchwork)
```

## Conceptual framing

Hidden Markov Models (HMMs) assume that the *observable* sequence of apps
is generated by an underlying sequence of *latent states* that we cannot
directly see. Each state has a characteristic probability distribution over
apps (emission probabilities) and transitions between states follow a Markov
process (transition matrix).

The goal is to learn: **how many latent behavioral states exist, what apps
each state is associated with, and how people move between states.**

The reference study (Fan et al., 2026) used a categorical HMM
(`hmmlearn.CategoricalHMM` in Python) on person-day sequences. We replicate
this using `depmixS4` in R, which fits multinomial HMMs to variable-length
sequences.

**Unit of analysis:** person-day sequences (same as reference) — i.e., all
app uses for one person on one day form one sequence. This keeps sequences
short enough for the EM algorithm to converge well, and aligns the model
with daily behavioral rhythms.

Alternative: phone-session sequences (using `phone_session_id`), which are
even shorter and capture single-pickup patterns. Both are implemented below.

---

## Data

```{r load-data}
# --- Real data (uncomment when available) ---
# app_categories <- read_csv("data/raw/app_categories.csv")  # provided by project
#
# df_raw <- read_csv(
#   "data/raw/uni_mainz_app_session.csv",
#   col_types = cols(
#     participant_code = col_character(),
#     start_time       = col_datetime(),
#     Duration         = col_double(),
#     phone_session_id = col_double(),
#     package_name     = col_character(),
#     app_name         = col_character(),
#     .default         = col_skip()
#   )
# ) |>
#   left_join(app_categories, by = "package_name") |>
#   arrange(participant_code, start_time)

# --- SYNTHETIC SAMPLE for development ---
set.seed(42)

pkg_to_app <- c(
  "com.instagram.android"         = "Instagram",
  "com.whatsapp"                  = "WhatsApp",
  "com.zhiliaoapp.musically"      = "TikTok",
  "com.google.android.youtube"    = "YouTube",
  "com.android.chrome"            = "Chrome",
  "com.google.android.apps.maps"  = "Maps",
  "com.google.android.calculator" = "Rechner",
  "com.google.android.gm"         = "Gmail",
  "com.facebook.orca"             = "Messenger",
  "com.spotify.music"             = "Spotify"
)
pkg_to_cat <- c(
  "com.instagram.android"         = "social network",
  "com.whatsapp"                  = "communication",
  "com.zhiliaoapp.musically"      = "social network",
  "com.google.android.youtube"    = "social network",
  "com.android.chrome"            = "tools",
  "com.google.android.apps.maps"  = "tools",
  "com.google.android.calculator" = "tools",
  "com.google.android.gm"         = "communication",
  "com.facebook.orca"             = "communication",
  "com.spotify.music"             = "entertainment"
)
pkg_names <- names(pkg_to_app)
pkg_probs  <- c(3, 3, 2, 2, 1, 1, .5, 1, 1, 1)

n_persons <- 60
events_per <- sample(100:400, n_persons, replace = TRUE)

df_raw <- map2_dfr(
  paste0("P", sprintf("%03d", seq_len(n_persons))),
  events_per,
  function(pid, n) {
    starts <- as.POSIXct("2024-04-01 08:00:00") +
      cumsum(c(0, sample(5:1800, n - 1, replace = TRUE)))
    pkgs <- sample(pkg_names, n, replace = TRUE, prob = pkg_probs)
    durs <- sample(5:900, n, replace = TRUE) + runif(n)
    tibble(
      participant_code = pid,
      package_name     = pkgs,
      app_name         = pkg_to_app[pkgs],
      app_category     = pkg_to_cat[pkgs],
      start_time       = starts,
      Duration         = durs,
      phone_session_id = cumsum(c(0, as.numeric(diff(starts)) > 300)) %% 100
    )
  }
) |> arrange(participant_code, start_time)
# --- end synthetic sample ---
```

---

## Data preparation

### Step 1 — Merge consecutive same-app episodes

```{r merge}
gap_threshold <- 10 * 60  # seconds

merge_consecutive <- function(d, gap_sec) {
  d <- arrange(d, start_time)
  out <- list(); cur <- d[1, ]
  for (i in seq_len(nrow(d))[-1]) {
    row <- d[i, ]
    gap <- as.numeric(row$start_time - (cur$start_time + cur$Duration), units = "secs")
    if (row$app_name == cur$app_name && !is.na(gap) && gap <= gap_sec) {
      cur$Duration <- cur$Duration + row$Duration
    } else {
      out <- c(out, list(cur)); cur <- row
    }
  }
  bind_rows(c(out, list(cur)))
}

df_merged <- df_raw |>
  filter(Duration > 0) |>
  group_by(participant_code) |>
  group_modify(~ merge_consecutive(.x, gap_threshold)) |>
  ungroup()
```

### Step 2 — Encode app_name as integer

`depmixS4` requires the categorical response to be an integer starting at 1.

```{r encode}
app_levels <- sort(unique(df_merged$app_name))
n_symbols  <- length(app_levels)

df_merged <- df_merged |>
  mutate(app_code = as.integer(factor(app_name, levels = app_levels)))

cat("Observable symbols (n =", n_symbols, "):\n")
tibble(code = seq_along(app_levels), app = app_levels) |> print()
```

### Step 3 — Define sequences: person-day

Split each person's data into daily sequences. The day boundary is derived
from `start_time`. Sequences shorter than `min_len` are dropped (too little
information for the HMM).

```{r person-day-seqs}
min_len <- 5   # minimum episodes per sequence (same as reference study)
max_len <- 50  # cap very long sequences to avoid domination; DECISION POINT

df_seqs <- df_merged |>
  mutate(date = as.Date(start_time)) |>
  group_by(participant_code, date) |>
  mutate(seq_len = n()) |>
  filter(seq_len >= min_len, seq_len <= max_len) |>
  ungroup()

seq_lengths <- df_seqs |>
  count(participant_code, date, name = "len")

seq_lengths |>
  pull(len) |>
  summary() |>
  print()

cat(sprintf(
  "\n%d person-day sequences retained (from %d persons)\n",
  nrow(seq_lengths), n_distinct(seq_lengths$participant_code)
))
```

---

## Model selection: AIC / BIC over k states

Fit categorical HMMs for k = 2 … 6 states. For each k, run multiple random
restarts and keep the best fit by log-likelihood (mirrors the reference approach).
Use AIC and BIC to choose k; prefer BIC for parsimony.

```{r model-selection, cache=TRUE}
# Concatenate all sequences; track lengths for depmixS4
obs_concat  <- df_seqs |> arrange(participant_code, date, start_time) |> pull(app_code)
ntimes_vec  <- seq_lengths |> arrange(participant_code, date) |> pull(len)

n_restarts <- 5   # increase for real data
k_range    <- 2:6

fit_hmm <- function(k, obs, ntimes, n_starts = 5) {
  best_ll <- -Inf; best_mod <- NULL
  for (s in seq_len(n_starts)) {
    mod <- depmix(
      response  = obs ~ 1,
      data      = data.frame(obs = obs),
      nstates   = k,
      ntimes    = ntimes,
      family    = multinomial("identity")
    )
    tryCatch({
      fit <- fit(mod, verbose = FALSE)
      ll  <- logLik(fit)
      if (ll > best_ll) { best_ll <- ll; best_mod <- fit }
    }, error = function(e) NULL)
  }
  best_mod
}

# Cache path — skip refitting if already computed
cache_path <- "results/tables/hmm_models.rds"
if (file.exists(cache_path)) {
  hmm_fits <- readRDS(cache_path)
} else {
  set.seed(99)
  hmm_fits <- lapply(setNames(k_range, k_range), \(k)
    fit_hmm(k, obs_concat, ntimes_vec, n_restarts))
  saveRDS(hmm_fits, cache_path)
}

# Collect AIC / BIC / log-likelihood
model_stats <- map_dfr(k_range, function(k) {
  m <- hmm_fits[[as.character(k)]]
  if (is.null(m)) return(NULL)
  tibble(
    k   = k,
    ll  = as.numeric(logLik(m)),
    aic = AIC(m),
    bic = BIC(m)
  )
})

model_stats |> knitr::kable(digits = 1, caption = "HMM model selection criteria")
```

```{r model-selection-plot, fig.width=8, fig.height=4}
model_stats |>
  pivot_longer(c(aic, bic), names_to = "criterion", values_to = "value") |>
  ggplot(aes(k, value, colour = toupper(criterion), group = criterion)) +
  geom_line() + geom_point(size = 2) +
  scale_colour_manual(values = c(AIC = "#4E79A7", BIC = "#59A14F")) +
  labs(title = "HMM model selection: AIC and BIC by number of states",
       x = "Number of hidden states (k)", y = "Criterion value (lower = better)",
       colour = NULL) +
  theme_minimal()
```

```{r choose-k}
# DECISION POINT: choose k based on the elbow in BIC.
# The reference used k = 4. Adjust after inspecting the plot above.
k_chosen <- model_stats |> slice_min(bic, n = 1) |> pull(k)
cat("BIC-optimal k:", k_chosen, "\n")

best_model <- hmm_fits[[as.character(k_chosen)]]
cat("Converged:", best_model@posterior |> nrow() > 0, "\n")
```

---

## Inspect the fitted model

### Transition matrix

```{r transition-matrix}
# Each best_model@transition[[s]] holds transition probabilities FROM state s.
# getpars() returns those probabilities as a vector of length k.
trans_mat <- t(sapply(seq_len(k_chosen), \(s) {
  getpars(best_model@transition[[s]])
}))
dimnames(trans_mat) <- list(
  paste0("From S", seq_len(k_chosen)),
  paste0("To S",   seq_len(k_chosen))
)

round(trans_mat, 3) |>
  as.data.frame() |>
  knitr::kable(caption = "Transition matrix (rows = from, cols = to)")
```

```{r transition-heatmap, fig.width=5, fig.height=4}
as.data.frame(trans_mat) |>
  rownames_to_column("from") |>
  pivot_longer(-from, names_to = "to", values_to = "prob") |>
  ggplot(aes(to, from, fill = prob)) +
  geom_tile(colour = "white", linewidth = 1.2) +
  geom_text(aes(label = sprintf("%.2f", prob)), size = 4) +
  scale_fill_distiller(palette = "Reds", direction = 1, limits = c(0, 1)) +
  labs(title = "HMM transition matrix", x = NULL, y = NULL, fill = "P") +
  theme_minimal() +
  theme(axis.text = element_text(size = 11))
```

### Emission probabilities

```{r emissions}
# Extract emission probabilities for each state
get_emissions <- function(model, k, symbols) {
  map_dfr(seq_len(k), function(s) {
    probs <- getpars(model@response[[s]][[1]])
    # depmixS4 multinomial response: parameters are the category probabilities
    tibble(
      state    = paste0("State ", s),
      app_name = symbols,
      prob     = probs[seq_along(symbols)]
    )
  })
}

emissions <- get_emissions(best_model, k_chosen, app_levels)

# Dominant app per state
emissions |>
  group_by(state) |>
  slice_max(prob, n = 3) |>
  arrange(state, desc(prob)) |>
  knitr::kable(digits = 3, caption = "Top 3 apps per latent state")
```

```{r emission-plot, fig.width=10, fig.height=5}
# Colour by app category for interpretability
cat_colours <- c(
  "social network" = "#F28E2B",
  "communication"  = "#4E79A7",
  "tools"          = "#59A14F",
  "entertainment"  = "#B07AA1",
  "other"          = "#BAB0AC"
)
# Build app → category map from the data
app_to_cat <- df_merged |>
  distinct(app_name, app_category) |>
  deframe()

emissions |>
  mutate(
    app_category = app_to_cat[app_name],
    fill_col     = cat_colours[app_category]
  ) |>
  ggplot(aes(prob, reorder(app_name, prob), fill = app_category)) +
  geom_col() +
  facet_wrap(~state, scales = "free_x") +
  scale_fill_manual(values = cat_colours, na.value = "#BAB0AC") +
  labs(
    x = "Emission probability", y = NULL,
    title = "Emission probabilities per latent state",
    fill  = "Category"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom", strip.text = element_text(face = "bold"))
```

### Initial state probabilities

```{r init-probs}
init_probs <- getpars(best_model)[seq_len(k_chosen)]
tibble(state = paste0("State ", seq_len(k_chosen)), init_prob = init_probs) |>
  knitr::kable(digits = 3, caption = "Initial state probabilities")
```

---

## Viterbi decoding: assign states to observations

The Viterbi algorithm finds the most likely hidden state sequence for each
observed sequence. This lets us label every episode with a state and
visualise state trajectories over time.

```{r viterbi}
# posterior() returns most likely state at each time step
state_seq <- posterior(best_model)

df_decoded <- df_seqs |>
  arrange(participant_code, date, start_time) |>
  mutate(
    state      = paste0("State ", state_seq$state),
    state_prob = apply(state_seq[, -1], 1, max)  # max posterior prob
  )
```

```{r state-over-day, fig.width=10, fig.height=4}
# How does state distribution change over the hours of the day?
df_decoded |>
  mutate(hour = as.integer(format(start_time, "%H"))) |>
  count(hour, state) |>
  group_by(hour) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(hour, pct, fill = state)) +
  geom_col() +
  scale_x_continuous(breaks = seq(0, 23, 3)) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    x = "Hour of day", y = "Proportion of episodes",
    title = "Latent state distribution by hour of day",
    fill  = "State"
  ) +
  theme_minimal()
```

```{r state-by-person, fig.width=9, fig.height=5}
# How many episodes per state per person? (person-level state profile)
df_decoded |>
  count(participant_code, state) |>
  group_by(participant_code) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(pct, fill = state)) +
  geom_histogram(bins = 20) +
  facet_wrap(~state) +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  labs(
    x = "Proportion of person's episodes in state",
    y = "Number of persons",
    title = "Person-level state profiles"
  ) +
  theme_minimal()
```

---

## Phone-session alternative

If phone sessions are short (< 20 apps per pickup), running the HMM on
session sequences rather than daily sequences is also interpretable and
captures single-pickup behavioral patterns.

```{r session-seqs}
# DECISION POINT: compare session-based vs. day-based results.

session_lengths <- df_merged |>
  count(participant_code, phone_session_id, name = "len")

cat("Session length distribution:\n")
summary(session_lengths$len)

# Only run session HMM if sessions are short enough to be interpretable
if (median(session_lengths$len) <= 15) {
  df_sess_seqs <- df_merged |>
    group_by(participant_code, phone_session_id) |>
    filter(n() >= 3) |>   # minimum 3 episodes per session
    ungroup()

  obs_sess   <- df_sess_seqs |>
    arrange(participant_code, phone_session_id, start_time) |>
    pull(app_code)
  ntimes_sess <- session_lengths |>
    filter(len >= 3) |>
    arrange(participant_code, phone_session_id) |>
    pull(len)

  cat(sprintf(
    "\nSession HMM: %d sequences, %d total observations\n",
    length(ntimes_sess), sum(ntimes_sess)
  ))
  # Fit session HMM with chosen k (reuse same k for comparability)
  # set.seed(99)
  # session_model <- fit_hmm(k_chosen, obs_sess, ntimes_sess)
  # Compare emission profiles: do session states match day states?
} else {
  cat("Sessions are long — person-day approach is preferred.\n")
}
```

---

## Summary of key outputs

| Output | What it shows |
|--------|--------------|
| AIC / BIC plot | How many latent states are supported by the data |
| Transition matrix heatmap | How persistent each state is; which transitions are common |
| Emission bar charts | Which apps characterise each latent state |
| Initial state probs | Which state people are most likely to start in |
| State-by-hour plot | Daily rhythm of latent behavioral states |
| Person-level profiles | How heterogeneous state use is across participants |
| Session alternative | Whether pickup-level HMM gives different/complementary states |
