---
title: "Hidden Markov Models of App Usage"
subtitle: "Method 3 — depmixS4 · ICA 2026 Hackathon"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    self-contained: true
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(depmixS4)
library(scales)
library(patchwork)
source("utils.R")
```

## Conceptual framing

Hidden Markov Models (HMMs) assume that the *observable* sequence of apps
is generated by an underlying sequence of *latent states* that we cannot
directly see. Each state has a characteristic probability distribution over
observations (emission probabilities) and transitions between states follow a
Markov process (transition matrix).

**Primary analysis — category-level HMM.** The emission unit is `app_category`
(5 symbols: social network, communication, tools, entertainment, other). This is
the approach used in Fan et al. (2026) and is preferred for two reasons:

1. *Interpretability*: a state whose emission distribution peaks on "social
   network" is immediately readable as a "social browsing state". A state whose
   distribution spans 116 individual apps is not.
2. *Statistical efficiency*: estimating 5 emission probabilities per state
   requires far less data than estimating 116, and EM converges reliably on the
   full dataset without any sequence sampling.

**Secondary analysis — app-level HMM.** Run on a capped sample to remain
feasible. Useful for checking whether within-category distinctions matter, but
the primary inference is at the category level.

**Post-hoc app enrichment.** After Viterbi decoding on categories, we look at
which specific apps dominate each decoded state. This gives both the clean
latent structure (from category HMM) and the fine-grained app detail.

**Unit of analysis:** person-day sequences — all app uses for one person on one
calendar day. Aligns with the reference study and captures daily behavioral
rhythms.

---

## Data

```{r load-data}
df_raw <- read_csv(
  "../_files/d_apps_anonymized.csv",
  col_types = cols(
    participant_code = col_character(),
    start_time       = col_datetime(),
    end_time         = col_datetime(),
    duration         = col_double(),
    phone_session_id = col_double(),
    app_session_id   = col_double(),
    package_name     = col_character(),
    app_name         = col_character(),
    category         = col_character(),
    .default         = col_skip()
  )
) |>
  rename(app_category = category, Duration = duration) |>
  mutate(app_category = translate_category(app_category)) |>
  arrange(participant_code, start_time)
```

---

## Data preparation

### Step 1 — Merge consecutive same-app episodes

```{r merge}
df_merged <- df_raw |>
  filter(Duration > 0) |>
  group_by(participant_code) |>
  group_modify(~ merge_consecutive(.x, gap_threshold)) |>
  ungroup() |>
  filter(!app_name %in% noise_apps)  # remove launcher/home-screen artefacts (see utils.R)
```

### Step 2 — Encode categorical variables as integers

`depmixS4` requires the response to be an integer starting at 1.

```{r encode}
cat_levels <- sort(unique(df_merged$app_category))
app_levels <- sort(unique(df_merged$app_name))

cat(sprintf("Category symbols: %d | App symbols: %d\n",
            length(cat_levels), length(app_levels)))

df_merged <- df_merged |>
  mutate(
    cat_code = as.integer(factor(app_category, levels = cat_levels)),
    app_code = as.integer(factor(app_name,     levels = app_levels))
  )
```

### Step 3 — Person-day sequences

```{r person-day-seqs}
min_len <- 5   # minimum episodes per day
max_len <- 25  # truncate long days (not discard — median is ~128 events/day)

df_seqs <- df_merged |>
  mutate(date = as.Date(start_time)) |>
  group_by(participant_code, date) |>
  mutate(
    len_before_cap = n(),       # raw daily count before truncation
    turn_in_day    = row_number()
  ) |>
  filter(len_before_cap >= min_len, turn_in_day <= max_len) |>
  ungroup()

seq_lengths <- df_seqs |>
  count(participant_code, date, name = "len")

raw_day_lengths <- df_seqs |>
  distinct(participant_code, date, len_before_cap) |>
  pull(len_before_cap)

cat(sprintf(
  "%d person-day sequences retained (from %d persons); lengths %d–%d\n",
  nrow(seq_lengths), n_distinct(seq_lengths$participant_code),
  min(seq_lengths$len), max(seq_lengths$len)
))
cat(sprintf(
  "Note: sequences truncated at turn %d; median day has %d events so ~%.0f%% of daily activity is modelled.\n",
  max_len, round(median(raw_day_lengths)), 100 * max_len / median(raw_day_lengths)
))
```

---

## Primary analysis: Category-level HMM

### Shared fitting helper

```{r fit-helper}
fit_hmm <- function(k, obs, ntimes, n_starts, n_symbols) {
  best_ll <- -Inf; best_mod <- NULL
  for (s in seq_len(n_starts)) {
    mod <- depmix(
      response = obs ~ 1,
      data     = data.frame(obs = obs),
      nstates  = k,
      ntimes   = ntimes,
      family   = multinomial("identity")
    )
    tryCatch({
      fit <- fit(mod, verbose = FALSE)
      ll  <- logLik(fit)
      if (ll > best_ll) { best_ll <- ll; best_mod <- fit }
    }, error = function(e) NULL)
  }
  best_mod
}

get_emissions <- function(model, k, symbols, col_name = "symbol") {
  map_dfr(seq_len(k), function(s) {
    probs <- getpars(model@response[[s]][[1]])
    tibble(
      state          = paste0("State ", s),
      !!col_name    := symbols,
      prob           = probs[seq_along(symbols)]
    )
  })
}
```

### Model selection

```{r cat-model-selection}
# Category-level: 5 emission symbols → no sequence sampling needed.
# Scales to the full 1000-person dataset without modification.
obs_cat    <- df_seqs |> arrange(participant_code, date, start_time) |> pull(cat_code)
ntimes_cat <- seq_lengths |> arrange(participant_code, date) |> pull(len)

n_restarts_cat <- 2
k_range_cat    <- 2:4

cat(sprintf("Category HMM training: %d sequences, %d total observations\n",
            nrow(seq_lengths), length(obs_cat)))

cache_cat <- "results/tables/hmm_cat_models.rds"
if (file.exists(cache_cat)) {
  hmm_cat_fits <- readRDS(cache_cat)
  # Invalidate if trained on different data
  if (!identical(attr(hmm_cat_fits, "ntimes"), ntimes_cat)) {
    hmm_cat_fits <- NULL
  }
}
if (!exists("hmm_cat_fits") || is.null(hmm_cat_fits)) {
  set.seed(99)
  hmm_cat_fits <- lapply(setNames(k_range_cat, k_range_cat), \(k)
    fit_hmm(k, obs_cat, ntimes_cat, n_restarts_cat, length(cat_levels)))
  attr(hmm_cat_fits, "ntimes") <- ntimes_cat
  saveRDS(hmm_cat_fits, cache_cat)
}

cat_stats <- map_dfr(k_range_cat, function(k) {
  m <- hmm_cat_fits[[as.character(k)]]
  if (is.null(m)) return(NULL)
  tibble(k = k, ll = as.numeric(logLik(m)), aic = AIC(m), bic = BIC(m))
})

cat_stats |> knitr::kable(digits = 1, caption = "Category HMM: model selection criteria")
```

```{r cat-model-selection-plot, fig.width=8, fig.height=4}
cat_stats |>
  pivot_longer(c(aic, bic), names_to = "criterion", values_to = "value") |>
  ggplot(aes(k, value, colour = toupper(criterion), group = criterion)) +
  geom_line() + geom_point(size = 2) +
  scale_colour_manual(values = c(AIC = "#4E79A7", BIC = "#59A14F")) +
  labs(title = "Category HMM: AIC and BIC by number of states",
       x = "Number of hidden states (k)", y = "Criterion value (lower = better)",
       colour = NULL) +
  theme_minimal()
```

```{r cat-choose-k}
k_cat <- cat_stats |> slice_min(bic, n = 1) |> pull(k)
cat("BIC-optimal k (category HMM):", k_cat, "\n")
best_cat_model <- hmm_cat_fits[[as.character(k_cat)]]
```

### Transition matrix

```{r cat-transition-matrix}
trans_mat <- t(sapply(seq_len(k_cat), \(s) getpars(best_cat_model@transition[[s]])))
dimnames(trans_mat) <- list(paste0("From S", seq_len(k_cat)),
                            paste0("To S",   seq_len(k_cat)))

round(trans_mat, 3) |>
  as.data.frame() |>
  knitr::kable(caption = "Transition matrix (rows = from, cols = to)")
```

```{r cat-transition-heatmap, fig.width=5, fig.height=4}
as.data.frame(trans_mat) |>
  rownames_to_column("from") |>
  pivot_longer(-from, names_to = "to", values_to = "prob") |>
  ggplot(aes(to, from, fill = prob)) +
  geom_tile(colour = "white", linewidth = 1.2) +
  geom_text(aes(label = sprintf("%.2f", prob)), size = 4) +
  scale_fill_distiller(palette = "Reds", direction = 1, limits = c(0, 1)) +
  labs(title = "HMM transition matrix", x = NULL, y = NULL, fill = "P") +
  theme_minimal() +
  theme(axis.text = element_text(size = 11))
```

### Emission probabilities

Each state's emission distribution over the 5 app categories directly names
what the state *is* — a peak on "social network" = a social browsing state, etc.

```{r cat-emissions}
emissions_cat <- get_emissions(best_cat_model, k_cat, cat_levels, col_name = "category")

emissions_cat |>
  group_by(state) |>
  slice_max(prob, n = 2) |>
  arrange(state, desc(prob)) |>
  knitr::kable(digits = 3, caption = "Dominant categories per latent state")
```

```{r cat-emission-plot, fig.width=9, fig.height=4}
emissions_cat |>
  ggplot(aes(prob, category, fill = category)) +
  geom_col() +
  facet_wrap(~state, nrow = 1) +
  scale_fill_manual(values = cat_pal, guide = "none") +
  scale_x_continuous(labels = percent_format()) +
  labs(x = "Emission probability", y = NULL,
       title = "Category emission probabilities per latent state",
       subtitle = "The dominant category names the state") +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"))
```

### Initial state probabilities

```{r cat-init-probs}
init_probs <- getpars(best_cat_model)[seq_len(k_cat)]
tibble(state = paste0("State ", seq_len(k_cat)), init_prob = init_probs) |>
  knitr::kable(digits = 3, caption = "Initial state probabilities")
```

### Viterbi decoding

```{r cat-viterbi}
state_seq <- posterior(best_cat_model)

df_decoded <- df_seqs |>
  arrange(participant_code, date, start_time) |>
  mutate(
    state      = paste0("State ", state_seq$state),
    state_prob = apply(state_seq[, -1], 1, max)
  )
```

### Post-hoc app enrichment

Viterbi decodes states from *categories*. Looking at which specific apps
appear most within each decoded state gives finer-grained interpretation
without sacrificing the clean latent structure.

```{r app-enrichment}
df_decoded |>
  count(state, app_name, sort = TRUE) |>
  group_by(state) |>
  mutate(pct = n / sum(n)) |>
  slice_max(pct, n = 5) |>
  ungroup() |>
  arrange(state, desc(pct)) |>
  dplyr::select(state, app_name, pct) |>
  knitr::kable(digits = 3,
               caption = "Top 5 apps within each decoded state (post-hoc enrichment)")
```

### State distribution by hour of day

```{r cat-state-over-day, fig.width=10, fig.height=4}
df_decoded |>
  mutate(hour = as.integer(format(start_time, "%H"))) |>
  count(hour, state) |>
  group_by(hour) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(hour, pct, fill = state)) +
  geom_col() +
  scale_x_continuous(breaks = seq(0, 23, 3)) +
  scale_fill_brewer(palette = "Set2") +
  labs(x = "Hour of day", y = "Proportion of episodes",
       title = "Latent state distribution by hour of day",
       fill  = "State") +
  theme_minimal()
```

### Person-level state profiles

```{r cat-state-by-person, fig.width=9, fig.height=4}
df_decoded |>
  count(participant_code, state) |>
  group_by(participant_code) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(x = state, y = pct, fill = state)) +
  geom_col() +
  facet_wrap(~participant_code, labeller = label_both) +
  scale_fill_brewer(palette = "Set2", guide = "none") +
  scale_y_continuous(labels = percent_format()) +
  labs(x = NULL, y = "Share of episodes",
       title = "Person-level state profiles") +
  theme_minimal()
```

---

## Secondary: App-level HMM

Run on a capped sample due to the large emission alphabet (116+ symbols).
Use to check whether within-category app distinctions create distinct states;
compare with the category-level result.

```{r app-model-selection}
# Cap at max_hmm_seqs sequences — app-level EM is slow with 116 symbols.
# DECISION POINT: raise to 300+ on full data (expect ~10 min per k per restart).
max_hmm_seqs <- 50
n_restarts_app <- 3   # DECISION POINT — raise to 5+ on full data
k_range_app    <- 2:4

set.seed(99)
hmm_days <- if (nrow(seq_lengths) > max_hmm_seqs) {
  slice_sample(seq_lengths, n = max_hmm_seqs)
} else {
  seq_lengths
}

df_hmm_app <- df_seqs |> semi_join(hmm_days, by = c("participant_code", "date"))
obs_app    <- df_hmm_app |> arrange(participant_code, date, start_time) |> pull(app_code)
ntimes_app <- hmm_days  |> arrange(participant_code, date) |> pull(len)

cat(sprintf("App HMM training: %d sequences, %d total observations\n",
            nrow(hmm_days), length(obs_app)))

cache_app <- "results/tables/hmm_app_models.rds"
if (file.exists(cache_app)) {
  hmm_app_fits <- readRDS(cache_app)
  if (!identical(attr(hmm_app_fits, "ntimes"), ntimes_app)) hmm_app_fits <- NULL
}
if (!exists("hmm_app_fits") || is.null(hmm_app_fits)) {
  set.seed(99)
  hmm_app_fits <- lapply(setNames(k_range_app, k_range_app), \(k)
    fit_hmm(k, obs_app, ntimes_app, n_restarts_app, length(app_levels)))
  attr(hmm_app_fits, "ntimes") <- ntimes_app
  saveRDS(hmm_app_fits, cache_app)
}

app_stats <- map_dfr(k_range_app, function(k) {
  m <- hmm_app_fits[[as.character(k)]]
  if (is.null(m)) return(NULL)
  tibble(k = k, ll = as.numeric(logLik(m)), aic = AIC(m), bic = BIC(m))
})

app_stats |> knitr::kable(digits = 1, caption = "App HMM: model selection criteria")
```

```{r app-emissions}
k_app <- app_stats |> slice_min(bic, n = 1) |> pull(k)
best_app_model <- hmm_app_fits[[as.character(k_app)]]

if (!is.null(best_app_model)) {
  app_to_cat <- df_merged |> distinct(app_name, app_category) |> deframe()

  get_emissions(best_app_model, k_app, app_levels, col_name = "app_name") |>
    mutate(app_category = app_to_cat[app_name]) |>
    group_by(state) |>
    slice_max(prob, n = 5) |>
    arrange(state, desc(prob)) |>
    dplyr::select(state, app_name, app_category, prob) |>
    knitr::kable(digits = 3, caption = "Top 5 apps per latent state (app-level HMM)")
}
```

---

## Phone-session alternative

```{r session-seqs}
session_lengths <- df_merged |>
  count(participant_code, phone_session_id, name = "len")

cat("Session length distribution:\n")
print(summary(session_lengths$len))

# Cap sessions for fitting — with 1000 participants this can reach 100k+ sessions.
max_sess_seqs <- 300  # DECISION POINT — sufficient for stable EM at category level
min_sess_len  <- 3

sess_eligible <- session_lengths |> filter(len >= min_sess_len)

set.seed(99)
sess_sample <- if (nrow(sess_eligible) > max_sess_seqs) {
  slice_sample(sess_eligible, n = max_sess_seqs)
} else {
  sess_eligible
}

df_sess <- df_merged |>
  semi_join(sess_sample, by = c("participant_code", "phone_session_id"))

obs_sess    <- df_sess |>
  arrange(participant_code, phone_session_id, start_time) |>
  pull(cat_code)
ntimes_sess <- sess_sample |>
  arrange(participant_code, phone_session_id) |>
  pull(len)

cat(sprintf("\nSession HMM: %d sequences, %d total observations\n",
            nrow(sess_sample), sum(ntimes_sess)))

set.seed(99)
session_model <- fit_hmm(k_cat, obs_sess, ntimes_sess, n_restarts_cat,
                         length(cat_levels))

if (!is.null(session_model)) {
  get_emissions(session_model, k_cat, cat_levels, col_name = "category") |>
    group_by(state) |>
    slice_max(prob, n = 2) |>
    arrange(state, desc(prob)) |>
    knitr::kable(digits = 3,
                 caption = "Dominant categories per state (session-based HMM)")
} else {
  message("Session HMM did not converge — try more restarts or a smaller k.")
}
```

---

## Summary of key outputs

| Output | What it shows |
|--------|--------------|
| AIC / BIC plot (category) | How many latent states are supported |
| Transition matrix heatmap | State persistence and common switches |
| Category emission plot | What each state *is* — directly readable |
| Initial state probs | Which state people typically start in |
| App enrichment table | Which specific apps dominate each decoded state |
| State-by-hour plot | Daily rhythm of latent behavioral states |
| Person-level profiles | How each person distributes across states |
| App-level HMM | Whether within-category distinctions create additional states |
| Session HMM | Whether pickup-level patterns differ from daily patterns |
