---
title: "Network Analysis of App Transitions"
subtitle: "Method 4 — igraph / ggraph · ICA 2026 Hackathon"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    self-contained: true
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(igraph)
library(tidygraph)
library(ggraph)
library(scales)
```

## Conceptual framing

We represent app-switching behaviour as a **directed weighted network**:

- **Nodes** = apps (one per unique `app_name`)
- **Directed edges** = observed transitions (app A immediately followed by app B)
- **Edge weights** = count of that transition across all persons and time

This directly mirrors Fan et al. (2026), who built the network from bigrams of
activity-type sequences. We use consecutive app pairs from the merged sequences.

Self-loops (same app → same app) are excluded — they are already compressed by
the merging step, and within-app repetition is not informative for switching.

Two network levels:
1. **App-level** — fine-grained, nodes coloured by `app_category`
2. **Category-level** — coarser, shows cross-category switching structure

---

## Data

```{r load-data}
# --- Real data (uncomment when available) ---
# app_categories <- read_csv("data/raw/app_categories.csv")  # provided by project
#
# df_raw <- read_csv(
#   "data/raw/uni_mainz_app_session.csv",
#   col_types = cols(
#     participant_code = col_character(),
#     start_time       = col_datetime(),
#     Duration         = col_double(),
#     phone_session_id = col_double(),
#     package_name     = col_character(),
#     app_name         = col_character(),
#     .default         = col_skip()
#   )
# ) |>
#   left_join(app_categories, by = "package_name") |>
#   arrange(participant_code, start_time)

# --- SYNTHETIC SAMPLE for development ---
set.seed(42)
pkg_to_app <- c(
  "com.instagram.android"         = "Instagram",
  "com.whatsapp"                  = "WhatsApp",
  "com.zhiliaoapp.musically"      = "TikTok",
  "com.google.android.youtube"    = "YouTube",
  "com.android.chrome"            = "Chrome",
  "com.google.android.apps.maps"  = "Maps",
  "com.google.android.calculator" = "Rechner",
  "com.google.android.gm"         = "Gmail",
  "com.facebook.orca"             = "Messenger",
  "com.spotify.music"             = "Spotify"
)
pkg_to_cat <- c(
  "com.instagram.android"         = "social network",
  "com.whatsapp"                  = "communication",
  "com.zhiliaoapp.musically"      = "social network",
  "com.google.android.youtube"    = "social network",
  "com.android.chrome"            = "tools",
  "com.google.android.apps.maps"  = "tools",
  "com.google.android.calculator" = "tools",
  "com.google.android.gm"         = "communication",
  "com.facebook.orca"             = "communication",
  "com.spotify.music"             = "entertainment"
)
pkg_names <- names(pkg_to_app)
pkg_probs  <- c(3, 3, 2, 2, 1, 1, .5, 1, 1, 1)

df_raw <- map2_dfr(
  paste0("P", sprintf("%03d", 1:80)),
  sample(100:500, 80, replace = TRUE),
  function(pid, n) {
    starts <- as.POSIXct("2024-04-01 08:00:00") +
      cumsum(c(0, sample(5:1800, n - 1, replace = TRUE)))
    pkgs <- sample(pkg_names, n, replace = TRUE, prob = pkg_probs)
    tibble(
      participant_code = pid,
      package_name     = pkgs,
      app_name         = pkg_to_app[pkgs],
      app_category     = pkg_to_cat[pkgs],
      start_time       = starts,
      Duration         = sample(5:900, n, replace = TRUE) + runif(n),
      phone_session_id = cumsum(c(0, as.numeric(diff(starts)) > 300)) %% 100
    )
  }
) |> arrange(participant_code, start_time)
# --- end synthetic sample ---
```

---

## Data preparation

### Merge consecutive same-app episodes

```{r merge}
gap_threshold <- 10 * 60

merge_consecutive <- function(d, gap_sec) {
  d <- arrange(d, start_time)
  out <- list(); cur <- d[1, ]
  for (i in seq_len(nrow(d))[-1]) {
    row <- d[i, ]
    gap <- as.numeric(row$start_time - (cur$start_time + cur$Duration), units = "secs")
    if (row$app_name == cur$app_name && !is.na(gap) && gap <= gap_sec) {
      cur$Duration <- cur$Duration + row$Duration
    } else {
      out <- c(out, list(cur)); cur <- row
    }
  }
  bind_rows(c(out, list(cur)))
}

df_merged <- df_raw |>
  filter(Duration > 0) |>
  group_by(participant_code) |>
  group_modify(~ merge_consecutive(.x, gap_threshold)) |>
  ungroup()
```

### Build edge list from consecutive app pairs

```{r edge-list}
# For the app-level network: transitions between app_name values
edges_app <- df_merged |>
  arrange(participant_code, start_time) |>
  group_by(participant_code) |>
  mutate(next_app = lead(app_name)) |>
  ungroup() |>
  filter(!is.na(next_app), app_name != next_app) |>   # remove self-loops
  count(from = app_name, to = next_app, name = "weight") |>
  arrange(desc(weight))

# For the category-level network
edges_cat <- df_merged |>
  arrange(participant_code, start_time) |>
  group_by(participant_code) |>
  mutate(next_cat = lead(app_category)) |>
  ungroup() |>
  filter(!is.na(next_cat), app_category != next_cat) |>
  count(from = app_category, to = next_cat, name = "weight") |>
  arrange(desc(weight))

cat("App-level edges (unique transitions):", nrow(edges_app), "\n")
cat("Category-level edges:                ", nrow(edges_cat), "\n")

head(edges_app, 10) |>
  knitr::kable(caption = "Top 10 most frequent app transitions")
```

---

## App-level network

### Build graph

```{r build-graph-app}
# Node attributes: one row per app with category
nodes_app <- df_merged |>
  distinct(app_name, app_category) |>
  rename(name = app_name)

g_app <- tbl_graph(
  nodes = nodes_app,
  edges = edges_app,
  directed = TRUE
) |>
  activate(nodes) |>
  mutate(
    degree_total = centrality_degree(mode = "all"),
    degree_in    = centrality_degree(mode = "in"),
    degree_out   = centrality_degree(mode = "out"),
    closeness    = centrality_closeness(mode = "all"),
    betweenness  = centrality_betweenness(directed = TRUE, weights = NULL),
    eigenvector  = centrality_eigen(directed = FALSE)
  )

cat("Nodes:", igraph::vcount(g_app), "| Edges:", igraph::ecount(g_app), "\n")
cat("Is directed:", igraph::is_directed(g_app), "\n")
```

### Centrality summary

```{r centrality-table}
g_app |>
  activate(nodes) |>
  as_tibble() |>
  arrange(desc(degree_total)) |>
  select(app = name, category = app_category,
         degree = degree_total, in_deg = degree_in,
         out_deg = degree_out, betweenness, eigenvector) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "Node centrality measures (sorted by degree)")
```

```{r centrality-plot, fig.width=9, fig.height=4}
cat_pal <- c(
  "social network" = "#F28E2B",
  "communication"  = "#4E79A7",
  "tools"          = "#59A14F",
  "entertainment"  = "#B07AA1",
  "other"          = "#BAB0AC"
)

g_app |>
  activate(nodes) |>
  as_tibble() |>
  pivot_longer(c(degree_in, degree_out, betweenness, eigenvector),
               names_to = "measure", values_to = "value") |>
  mutate(measure = recode(measure,
    degree_in   = "In-degree",
    degree_out  = "Out-degree",
    betweenness = "Betweenness",
    eigenvector = "Eigenvector"
  )) |>
  ggplot(aes(value, reorder(name, value), fill = app_category)) +
  geom_col() +
  facet_wrap(~measure, scales = "free_x") +
  scale_fill_manual(values = cat_pal, na.value = "#BAB0AC") +
  labs(x = NULL, y = NULL, fill = "Category",
       title = "Centrality measures by app") +
  theme_minimal() +
  theme(legend.position = "bottom", axis.text.y = element_text(size = 8))
```

### Visualisation

```{r network-viz-app, fig.width=10, fig.height=8}
# Threshold: only draw edges above a minimum weight for legibility
# DECISION POINT: adjust weight_threshold based on actual data volume
weight_threshold <- quantile(edges_app$weight, 0.5)

g_app |>
  activate(edges) |>
  filter(weight >= weight_threshold) |>
  ggraph(layout = "fr") +   # Fruchterman-Reingold ≈ ForceAtlas2 in reference
  geom_edge_arc(
    aes(width = weight, alpha = weight),
    colour    = "grey50",
    arrow     = arrow(length = unit(3, "mm"), type = "closed"),
    end_cap   = circle(4, "mm"),
    strength  = 0.15
  ) +
  geom_node_point(
    aes(size = degree_total, colour = app_category),
    alpha = 0.9
  ) +
  geom_node_text(
    aes(label = name),
    repel = TRUE, size = 3.2, colour = "grey20"
  ) +
  scale_edge_width(range = c(0.3, 3), guide = "none") +
  scale_edge_alpha(range = c(0.2, 0.8), guide = "none") +
  scale_size(range = c(3, 12), name = "Degree") +
  scale_colour_manual(values = cat_pal, name = "Category",
                      na.value = "#BAB0AC") +
  labs(title = "App transition network",
       subtitle = paste0("Edges ≥ ", round(weight_threshold),
                         " transitions; node size = degree")) +
  theme_graph(base_family = "sans") +
  theme(legend.position = "right")
```

### Community detection

```{r communities-app}
# Convert to undirected for modularity-based community detection
# (Louvain works on undirected; for directed use cluster_walktrap)
g_undirected <- g_app |>
  activate(edges) |>
  mutate(weight = weight) |>
  convert(to_undirected)

communities <- cluster_louvain(g_undirected, weights = E(g_undirected)$weight)

cat(sprintf(
  "Communities detected: %d  |  Modularity: %.3f\n",
  length(communities), modularity(communities)
))

# Assign community membership back to nodes
g_app <- g_app |>
  activate(nodes) |>
  mutate(community = as.character(membership(communities)))

g_app |>
  activate(nodes) |>
  as_tibble() |>
  select(app = name, category = app_category, community) |>
  arrange(community, app) |>
  knitr::kable(caption = "Community membership per app")
```

```{r community-viz, fig.width=10, fig.height=8}
g_app |>
  activate(edges) |>
  filter(weight >= weight_threshold) |>
  ggraph(layout = "fr") +
  geom_edge_arc(
    aes(width = weight, alpha = weight),
    colour   = "grey60",
    arrow    = arrow(length = unit(3, "mm"), type = "closed"),
    end_cap  = circle(4, "mm"),
    strength = 0.15
  ) +
  geom_node_point(aes(size = degree_total, colour = community), alpha = 0.9) +
  geom_node_text(aes(label = name), repel = TRUE, size = 3.2) +
  scale_edge_width(range = c(0.3, 3), guide = "none") +
  scale_edge_alpha(range = c(0.2, 0.8), guide = "none") +
  scale_size(range = c(3, 12), name = "Degree") +
  scale_colour_brewer(palette = "Set2", name = "Community") +
  labs(title = "App transition network — community structure") +
  theme_graph(base_family = "sans") +
  theme(legend.position = "right")
```

---

## Category-level network

Collapsing to categories reveals the cross-category switching structure
more clearly and is easier to interpret in a paper figure.

```{r build-graph-cat}
nodes_cat <- tibble(name = sort(unique(c(edges_cat$from, edges_cat$to))))

g_cat <- tbl_graph(nodes = nodes_cat, edges = edges_cat, directed = TRUE) |>
  activate(nodes) |>
  mutate(
    degree_total = centrality_degree(mode = "all"),
    strength_in  = centrality_degree(mode = "in",  weights = E(.)$weight),
    strength_out = centrality_degree(mode = "out", weights = E(.)$weight)
  )
```

```{r category-network-viz, fig.width=7, fig.height=6}
# Normalise edge weights to proportions for the chord-diagram feel
total_transitions <- sum(edges_cat$weight)

g_cat |>
  ggraph(layout = "circle") +
  geom_edge_arc(
    aes(width = weight / total_transitions * 100,
        alpha = weight,
        colour = from),
    arrow    = arrow(length = unit(4, "mm"), type = "closed"),
    end_cap  = circle(6, "mm"),
    strength = 0.3
  ) +
  geom_node_point(aes(colour = name, size = degree_total)) +
  geom_node_text(aes(label = name), vjust = -1.2, size = 4, fontface = "bold") +
  scale_edge_width(range = c(0.5, 5), name = "% of transitions") +
  scale_edge_alpha(range = c(0.3, 0.9), guide = "none") +
  scale_edge_colour_manual(values = cat_pal, guide = "none") +
  scale_colour_manual(values = cat_pal, name = "Category") +
  scale_size(range = c(5, 15), name = "Degree") +
  labs(title = "Category-level transition network") +
  theme_graph(base_family = "sans") +
  theme(legend.position = "right")
```

```{r category-transition-matrix}
# Transition rate matrix (row = from, col = to)
edges_cat |>
  group_by(from) |>
  mutate(rate = weight / sum(weight)) |>
  ungroup() |>
  select(from, to, rate) |>
  pivot_wider(names_from = to, values_from = rate, values_fill = 0) |>
  column_to_rownames("from") |>
  round(3) |>
  knitr::kable(caption = "Category transition rates (row sums to 1 within source category)")
```

---

## Within-session network

Building the network from within-session transitions only (no cross-session
edges) gives a cleaner view of app-switching within a single phone pickup,
uncontaminated by the natural break between sessions.

```{r within-session-network}
edges_session <- df_merged |>
  arrange(participant_code, phone_session_id, start_time) |>
  group_by(participant_code, phone_session_id) |>
  mutate(next_app = lead(app_name)) |>
  ungroup() |>
  filter(!is.na(next_app), app_name != next_app) |>
  count(from = app_name, to = next_app, name = "weight") |>
  arrange(desc(weight))

cat("Within-session transitions:\n")
head(edges_session, 10) |> knitr::kable()
```

```{r session-vs-full-comparison}
# What proportion of transitions are within-session vs. cross-session?
total_all     <- sum(edges_app$weight)
total_session <- sum(edges_session$weight)

cat(sprintf(
  "Within-session transitions: %d (%.1f%% of all transitions)\n",
  total_session, 100 * total_session / total_all
))

# Are the most common transitions the same within-session vs. overall?
left_join(
  edges_app     |> slice_max(weight, n = 10) |> mutate(rank_all     = row_number()),
  edges_session |> slice_max(weight, n = 10) |> mutate(rank_session = row_number()),
  by = c("from", "to")
) |>
  select(from, to, weight_all = weight.x, rank_all,
         weight_session = weight.y, rank_session) |>
  knitr::kable(caption = "Top 10 transitions: full data vs. within-session")
```

---

## Summary of key outputs

| Output | What it shows |
|--------|--------------|
| Edge list | Raw bigram counts; most common app-to-app transitions |
| Centrality table | Which apps are most central hubs in the switching network |
| Centrality bar charts | In/out-degree, betweenness, eigenvector per app |
| Force-directed network (app) | Full transition graph with node size = degree, colour = category |
| Community detection | Whether apps cluster into switching communities |
| Community network | Same graph coloured by community membership |
| Category network | Cross-category switching structure (cleaner poster figure) |
| Transition rate matrix | Conditional probability of switching from one category to another |
| Within-session network | Transitions within a single phone pickup only |
