---
title: "Process Mining of App Usage"
subtitle: "Method 5 — bupaR · ICA 2026 Hackathon"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    self-contained: true
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(bupaR)
library(processmapR)
library(heuristicsmineR)
library(scales)
library(patchwork)
```

## Conceptual framing

Process mining treats app usage as a **business process**: each person-day
is a *case*, each app use is an *event*, and the sequence of events forms a
*trace*. The goal is to discover recurring process patterns — frequent routes
through the app landscape — and to visualise them as process models.

The reference study (Fan et al., 2026) used `pm4py` in Python plus ProM Lite.
We replicate the key analyses using the `bupaR` ecosystem in R:

| Reference (pm4py / ProM) | This file (bupaR) |
|--------------------------|-------------------|
| `format_dataframe()` | `eventlog()` |
| `filter_variants_top_k()` | `filter_top_variants()` |
| `discover_dfg()` | `process_map(frequency())` |
| `discover_heuristics_net()` | `heuristics_map()` |
| `discover_bpmn_inductive()` | `process_map()` on filtered log |
| Event distribution plots | `dotted_chart()`, manual ggplot |

**Unit (case):** person-day — all app uses for one person on one calendar
day. This keeps traces short enough for process models to be readable.

**Activity:** `app_name` (fine-grained) or `app_category` (coarser).

**Extension vs. reference:** We add an explicit `app_switch` event each time
the person moves to a different app, making switches a first-class event in
the process model — mirroring the `platform_switch` synthetic event in the
reference.

---

## Data

```{r load-data}
# --- Real data (uncomment when available) ---
# app_categories <- read_csv("data/raw/app_categories.csv")
# df_raw <- read_csv(
#   "data/raw/uni_mainz_app_session.csv",
#   col_types = cols(
#     participant_code = col_character(),
#     start_time       = col_datetime(),
#     end_time         = col_datetime(),
#     Duration         = col_double(),
#     phone_session_id = col_double(),
#     package_name     = col_character(),
#     app_name         = col_character(),
#     .default         = col_skip()
#   )
# ) |>
#   left_join(app_categories, by = "package_name") |>
#   arrange(participant_code, start_time)

# --- SYNTHETIC SAMPLE ---
set.seed(42)
pkg_to_app <- c(
  "com.instagram.android" = "Instagram", "com.whatsapp" = "WhatsApp",
  "com.zhiliaoapp.musically" = "TikTok", "com.google.android.youtube" = "YouTube",
  "com.android.chrome" = "Chrome", "com.google.android.apps.maps" = "Maps",
  "com.google.android.calculator" = "Rechner", "com.google.android.gm" = "Gmail",
  "com.facebook.orca" = "Messenger", "com.spotify.music" = "Spotify"
)
pkg_to_cat <- c(
  "com.instagram.android" = "social network", "com.whatsapp" = "communication",
  "com.zhiliaoapp.musically" = "social network", "com.google.android.youtube" = "social network",
  "com.android.chrome" = "tools", "com.google.android.apps.maps" = "tools",
  "com.google.android.calculator" = "tools", "com.google.android.gm" = "communication",
  "com.facebook.orca" = "communication", "com.spotify.music" = "entertainment"
)
pkg_names  <- names(pkg_to_app)
pkg_probs  <- c(3, 3, 2, 2, 1, 1, .5, 1, 1, 1)

df_raw <- map2_dfr(
  paste0("P", sprintf("%03d", 1:60)),
  sample(80:300, 60, replace = TRUE),
  function(pid, n) {
    starts <- as.POSIXct("2024-04-01 08:00:00") +
      cumsum(c(0, sample(30:3600, n - 1, replace = TRUE)))
    pkgs <- sample(pkg_names, n, replace = TRUE, prob = pkg_probs)
    durs <- sample(10:900, n, replace = TRUE) + runif(n)
    tibble(
      participant_code = pid,
      package_name = pkgs, app_name = pkg_to_app[pkgs],
      app_category = pkg_to_cat[pkgs],
      start_time = starts, end_time = starts + durs, Duration = durs
    )
  }
) |> arrange(participant_code, start_time)
# --- end synthetic sample ---
```

---

## Data preparation

### Merge consecutive same-app episodes

```{r merge}
gap_threshold <- 10 * 60

merge_consecutive <- function(d, gap_sec) {
  d <- arrange(d, start_time); out <- list(); cur <- d[1, ]
  for (i in seq_len(nrow(d))[-1]) {
    row <- d[i, ]
    gap <- as.numeric(row$start_time - cur$end_time, units = "secs")
    if (row$app_name == cur$app_name && !is.na(gap) && gap <= gap_sec) {
      cur$end_time <- row$end_time; cur$Duration <- cur$Duration + row$Duration
    } else { out <- c(out, list(cur)); cur <- row }
  }
  bind_rows(c(out, list(cur)))
}

df_merged <- df_raw |>
  filter(Duration > 0) |>
  group_by(participant_code) |>
  group_modify(~ merge_consecutive(.x, gap_threshold)) |>
  ungroup()
```

### Build event log format

Cases are person-days. Each app-use episode is one event. We also inject a
synthetic `[app switch]` event at each transition between different apps,
mirroring the `platform_switch` event in the reference.

```{r build-events}
df_events <- df_merged |>
  mutate(date = as.Date(start_time)) |>
  # person-day case ID
  group_by(participant_code, date) |>
  mutate(
    day_rank = cur_group_id(),
    case_id  = paste0(participant_code, "_day", as.integer(date - min(date)) + 1),
    next_app = lead(app_name)
  ) |>
  ungroup()

# Inject [app switch] rows at each transition
switch_rows <- df_events |>
  filter(!is.na(next_app), app_name != next_app) |>
  mutate(
    app_name     = "[app switch]",
    app_category = "switch",
    # timestamp 1 second before the next event
    start_time   = end_time - 1,
    Duration     = NA_real_
  )

df_log_raw <- bind_rows(df_events, switch_rows) |>
  arrange(case_id, start_time) |>
  select(case_id, participant_code, app_name, app_category,
         start_time, end_time, Duration)

# Filter case lengths: 3 < events < 100 (mirrors reference filter)
case_lengths <- df_log_raw |> count(case_id, name = "n_events")
valid_cases  <- case_lengths |> filter(n_events > 3, n_events < 100) |> pull(case_id)

df_log_raw <- df_log_raw |> filter(case_id %in% valid_cases)

cat(sprintf(
  "Event log: %d cases, %d events, %d unique activities\n",
  n_distinct(df_log_raw$case_id),
  nrow(df_log_raw),
  n_distinct(df_log_raw$app_name)
))
```

### Create bupaR event log

```{r create-eventlog}
# bupaR requires: case_id, activity_id, activity_instance_id,
#                 lifecycle (start/complete), timestamp, resource
df_bupar <- df_log_raw |>
  mutate(
    # Each row is a complete activity instance
    activity_instance = row_number(),
    lifecycle         = "complete",
    resource          = app_category
  )

el <- eventlog(
  df_bupar,
  case_id              = "case_id",
  activity_id          = "app_name",
  activity_instance_id = "activity_instance",
  lifecycle_id         = "lifecycle",
  timestamp            = "start_time",
  resource_id          = "resource"
)

summary(el)
```

---

## Exploratory process analysis

```{r activity-frequency}
# Most frequent activities across all cases
el |>
  activity_frequency("activity") |>
  arrange(desc(absolute)) |>
  head(15) |>
  knitr::kable(caption = "Top 15 most frequent activities")
```

```{r case-length-dist, fig.width=7, fig.height=3}
el |>
  case_info() |>
  ggplot(aes(trace_length)) +
  geom_histogram(bins = 30, fill = "#4E79A7", colour = "white") +
  labs(x = "Events per case (person-day)", y = "Count",
       title = "Case length distribution") +
  theme_minimal()
```

```{r dotted-chart, fig.width=10, fig.height=6}
# Dotted chart: one row per case, dots = events, x = time within case
# Shows patterns across all cases at a glance
el |>
  dotted_chart(x = "relative", y = "start", color = "app_name") +
  labs(title = "Dotted chart: app events within each person-day case")
```

---

## Process maps (Directly Follows Graph)

```{r process-map-freq, fig.width=10, fig.height=7}
# Frequency-based process map — mirrors pm4py discover_dfg()
el |>
  process_map(
    type         = frequency("relative"),
    render       = TRUE
  )
```

```{r process-map-perf, fig.width=10, fig.height=7}
# Performance-based: median duration per activity
# Only meaningful for activities with Duration recorded (excludes [app switch])
el_timed <- df_bupar |>
  filter(app_name != "[app switch]", !is.na(Duration)) |>
  mutate(lifecycle = "complete") |>
  eventlog(
    case_id = "case_id", activity_id = "app_name",
    activity_instance_id = "activity_instance",
    lifecycle_id = "lifecycle", timestamp = "start_time",
    resource_id = "resource"
  )

el_timed |>
  process_map(
    type   = performance(mean, units = "secs"),
    render = TRUE
  )
```

### Filter to top-k variants

```{r top-variants}
# Keep the top 10 most frequent trace variants (mirrors filter_variants_top_k)
el_top10 <- el |> filter_top_variants(n = 10)

cat(sprintf(
  "Top 10 variants cover %d of %d cases (%.1f%%)\n",
  n_case(el_top10), n_case(el),
  100 * n_case(el_top10) / n_case(el)
))

# Show the variant list
el_top10 |>
  trace_coverage("case") |>
  head(10) |>
  knitr::kable(caption = "Top 10 most frequent trace variants")
```

```{r top-variants-map, fig.width=10, fig.height=7}
el_top10 |>
  process_map(type = frequency("relative"), render = TRUE)
```

---

## Heuristics miner

The heuristics miner is more noise-tolerant than the inductive miner and
produces a *heuristics net* showing the most reliable activity connections.

```{r heuristics-map, fig.width=10, fig.height=7}
# dependency_threshold: only keep edges with dependency score > threshold
# DECISION POINT: increase threshold to simplify the model
el |>
  heuristics_map(dependency_threshold = 0.5)
```

---

## Event distribution analysis

```{r hour-distribution, fig.width=8, fig.height=4}
df_log_raw |>
  filter(app_name != "[app switch]") |>
  mutate(hour = as.integer(format(start_time, "%H"))) |>
  count(hour, app_category) |>
  group_by(hour) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(hour, pct, fill = app_category)) +
  geom_col() +
  scale_x_continuous(breaks = seq(0, 23, 2)) +
  scale_fill_manual(values = c(
    "social network" = "#F28E2B", "communication" = "#4E79A7",
    "tools" = "#59A14F", "entertainment" = "#B07AA1",
    "switch" = "#E0E0E0", "other" = "#BAB0AC"
  )) +
  labs(x = "Hour of day", y = "Proportion of events",
       title = "App category usage by hour of day", fill = "Category") +
  theme_minimal()
```

```{r weekday-distribution, fig.width=8, fig.height=4}
df_log_raw |>
  filter(app_name != "[app switch]") |>
  mutate(weekday = weekdays(as.Date(start_time)),
         weekday = factor(weekday, levels = c(
           "Monday","Tuesday","Wednesday","Thursday","Friday","Saturday","Sunday"))) |>
  count(weekday, app_category) |>
  group_by(weekday) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(weekday, pct, fill = app_category)) +
  geom_col() +
  scale_fill_manual(values = c(
    "social network" = "#F28E2B", "communication" = "#4E79A7",
    "tools" = "#59A14F", "entertainment" = "#B07AA1",
    "switch" = "#E0E0E0", "other" = "#BAB0AC"
  )) +
  labs(x = NULL, y = "Proportion", fill = "Category",
       title = "App category usage by day of week") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

---

## Category transition duration heatmap

How long (median minutes) do people spend in each category before switching
to each other category? Mirrors the platform transition heatmap in the
reference (their `duration_time.csv`).

```{r transition-heatmap, fig.width=6, fig.height=5}
cat_pal_rocket <- c(
  "social network" = "#F28E2B", "communication" = "#4E79A7",
  "tools" = "#59A14F", "entertainment" = "#B07AA1"
)
core_cats <- names(cat_pal_rocket)

df_transitions <- df_merged |>
  filter(app_category %in% core_cats) |>
  arrange(participant_code, start_time) |>
  group_by(participant_code) |>
  mutate(
    next_cat = lead(app_category),
    dur_min  = Duration / 60
  ) |>
  ungroup() |>
  filter(!is.na(next_cat), app_category != next_cat,
         next_cat %in% core_cats)

heatmap_data <- df_transitions |>
  group_by(from = app_category, to = next_cat) |>
  summarise(median_dur_min = median(dur_min), .groups = "drop") |>
  complete(from = core_cats, to = core_cats, fill = list(median_dur_min = NA))

heatmap_data |>
  ggplot(aes(to, from, fill = median_dur_min)) +
  geom_tile(colour = "white", linewidth = 1) +
  geom_text(aes(label = ifelse(is.na(median_dur_min), "—",
                               sprintf("%.1f", median_dur_min))),
            size = 4) +
  scale_fill_distiller(palette = "Reds", direction = 1, na.value = "grey90",
                       name = "Median\nduration\n(min)") +
  labs(x = "To category", y = "From category",
       title = "Median duration before switching to each category") +
  theme_minimal() +
  theme(axis.text = element_text(size = 11),
        axis.title = element_text(size = 11))
```

---

## Category-level process map

A coarser, more interpretable process map using `app_category` as the
activity — cleaner for a paper figure than the app-level map.

```{r cat-eventlog}
df_bupar_cat <- df_log_raw |>
  filter(app_name != "[app switch]") |>
  mutate(
    activity_instance = row_number(),
    lifecycle         = "complete",
    resource          = "all"
  )

el_cat <- eventlog(
  df_bupar_cat,
  case_id              = "case_id",
  activity_id          = "app_category",
  activity_instance_id = "activity_instance",
  lifecycle_id         = "lifecycle",
  timestamp            = "start_time",
  resource_id          = "resource"
)
```

```{r cat-process-map, fig.width=8, fig.height=6}
el_cat |>
  process_map(type = frequency("relative"), render = TRUE)
```

---

## Summary of key outputs

| Output | What it shows |
|--------|--------------|
| Activity frequency | Which apps appear most in the event log |
| Case length distribution | How many app uses per person-day |
| Dotted chart | All events across all cases at a glance |
| Frequency process map | Most common app-to-app flows (= DFG) |
| Performance process map | Mean duration spent per app |
| Top-10 variant map | Most frequent full trace patterns |
| Heuristics net | Noise-tolerant process model |
| Hour / weekday distributions | Temporal rhythms of app use |
| Transition duration heatmap | How long in each category before switching |
| Category process map | Coarse-grained process model for the paper |
