---
title: "Process Mining of App Usage"
subtitle: "Method 5 — bupaR · ICA 2026 Hackathon"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    self-contained: true
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(bupaR)
library(processmapR)
library(scales)
library(patchwork)
source("utils.R")
```

## Conceptual framing

Process mining treats app usage as a **business process**: each person-day
is a *case*, each app use is an *event*, and the sequence of events forms a
*trace*. The goal is to discover recurring process patterns — frequent routes
through the app landscape — and to visualise them as process models.

The reference study (Fan et al., 2026) used `pm4py` in Python plus ProM Lite.
We replicate the key analyses using the `bupaR` ecosystem in R:

| Reference (pm4py / ProM) | This file (bupaR) |
|--------------------------|-------------------|
| `format_dataframe()` | `eventlog()` |
| `discover_dfg()` | `process_map(frequency())` on categories |
| Event distribution plots | `dotted_chart()`, manual ggplot |

**Unit (case):** person-day — all app uses for one person on one calendar
day. This keeps traces manageable.

**Activity:** `app_category` — app-level granularity produces unreadable
process maps with 100+ nodes; category-level gives interpretable models.

**Preprocessing:** direct app usage episodes — Pixel Launcher, Xperia Home, and
any other home-screen artefacts are excluded (see `utils.R`).

---

## Data

```{r load-data}
df_raw <- read_csv(
  "../_files/d_apps_anonymized.csv",
  col_types = cols(
    participant_code = col_character(),
    start_time       = col_datetime(),
    end_time         = col_datetime(),
    duration         = col_double(),
    phone_session_id = col_double(),
    app_session_id   = col_double(),
    package_name     = col_character(),
    app_name         = col_character(),
    category         = col_character(),
    .default         = col_skip()
  )
) |>
  rename(app_category = category, Duration = duration) |>
  mutate(app_category = translate_category(app_category)) |>
  apply_category_overrides() |>
  arrange(participant_code, start_time)
```

---

## Data preparation

### Merge consecutive same-app episodes

```{r merge}
df_merged <- df_raw |>
  filter(Duration > 0) |>
  group_by(participant_code) |>
  group_modify(~ merge_consecutive(.x, gap_threshold)) |>
  ungroup() |>
  filter(!app_name %in% noise_apps)
```

### Build event log format

Cases are person-days. Each app-use episode is one event.

```{r build-events}
df_events <- df_merged |>
  mutate(date = as.Date(start_time)) |>
  group_by(participant_code, date) |>
  mutate(case_id = paste0(participant_code, "_", date)) |>
  ungroup()

df_log_raw <- df_events |>
  arrange(case_id, start_time) |>
  select(case_id, participant_code, app_name, app_category,
         start_time, end_time, Duration)

# Filter case lengths: 3 < events < 1000
case_lengths <- df_log_raw |> count(case_id, name = "n_events")

cat("Case length distribution (app events per person-day):\n")
summary(case_lengths$n_events) |> print()

valid_cases <- case_lengths |> filter(n_events > 3, n_events < 1000) |> pull(case_id)
df_log_raw <- df_log_raw |> filter(case_id %in% valid_cases)

cat(sprintf(
  "Event log: %d cases, %d events, %d unique activities\n",
  n_distinct(df_log_raw$case_id),
  nrow(df_log_raw),
  n_distinct(df_log_raw$app_name)
))
```

### Create bupaR event log

```{r create-eventlog}
df_bupar <- df_log_raw |>
  mutate(
    activity_instance = row_number(),
    lifecycle         = "complete",
    resource          = app_category
  )

el <- eventlog(
  df_bupar,
  case_id              = "case_id",
  activity_id          = "app_name",
  activity_instance_id = "activity_instance",
  lifecycle_id         = "lifecycle",
  timestamp            = "start_time",
  resource_id          = "resource"
)

summary(el)
```

---

## Exploratory process analysis

```{r activity-frequency}
el |>
  activity_frequency("activity") |>
  arrange(desc(absolute)) |>
  head(15) |>
  knitr::kable(caption = "Top 15 most frequent activities")
```

```{r case-length-dist, fig.width=7, fig.height=3}
df_log_raw |>
  count(case_id, name = "trace_length") |>
  ggplot(aes(trace_length)) +
  geom_histogram(bins = 30, fill = "#4E79A7", colour = "white") +
  labs(x = "Events per case (person-day)", y = "Count",
       title = "Case length distribution") +
  theme_minimal()
```

```{r dotted-chart, fig.width=10, fig.height=6}
tryCatch(
  print(el |>
    dotted_chart(x = "relative", color = "resource") +
    scale_color_manual(values = cat_pal) +
    labs(title = "Dotted chart: app category events within each person-day case",
         colour = "Category")),
  error = function(e) message("dotted_chart: ", conditionMessage(e))
)
```

---

## Event distribution analysis

```{r hour-distribution, fig.width=8, fig.height=4}
df_log_raw |>
  mutate(hour = as.integer(format(start_time, "%H"))) |>
  count(hour, app_category) |>
  group_by(hour) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(hour, pct, fill = app_category)) +
  geom_col() +
  scale_x_continuous(breaks = seq(0, 23, 2)) +
  scale_fill_manual(values = cat_pal) +
  labs(x = "Hour of day", y = "Proportion of events",
       title = "App category usage by hour of day", fill = "Category") +
  theme_minimal()
```

```{r weekday-distribution, fig.width=8, fig.height=4}
df_log_raw |>
  mutate(weekday = weekdays(as.Date(start_time)),
         weekday = factor(weekday, levels = c(
           "Monday","Tuesday","Wednesday","Thursday","Friday","Saturday","Sunday"))) |>
  count(weekday, app_category) |>
  group_by(weekday) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(weekday, pct, fill = app_category)) +
  geom_col() +
  scale_fill_manual(values = cat_pal) +
  labs(x = NULL, y = "Proportion", fill = "Category",
       title = "App category usage by day of week") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

---

## Category transition duration heatmap

How long (median minutes) do people spend in each category before switching
to each other category?

```{r transition-heatmap, fig.width=6, fig.height=5}
cat_pal_rocket <- c(
  "communication"  = "#4E79A7", "social media"   = "#F28E2B",
  "entertainment"  = "#B07AA1", "gaming"         = "#E15759",
  "utilities"      = "#59A14F", "shopping"       = "#76B7B2",
  "practical life" = "#EDC948"
)
core_cats <- names(cat_pal_rocket)

df_transitions <- df_merged |>
  filter(app_category %in% core_cats) |>
  arrange(participant_code, start_time) |>
  group_by(participant_code) |>
  mutate(
    next_cat = lead(app_category),
    dur_min  = Duration / 60
  ) |>
  ungroup() |>
  filter(!is.na(next_cat), app_category != next_cat,
         next_cat %in% core_cats)

heatmap_data <- df_transitions |>
  group_by(from = app_category, to = next_cat) |>
  summarise(median_dur_min = median(dur_min), .groups = "drop") |>
  complete(from = core_cats, to = core_cats, fill = list(median_dur_min = NA))

heatmap_data |>
  ggplot(aes(to, from, fill = median_dur_min)) +
  geom_tile(colour = "white", linewidth = 1) +
  geom_text(aes(label = ifelse(is.na(median_dur_min), "—",
                               sprintf("%.1f", median_dur_min))),
            size = 4) +
  scale_fill_distiller(palette = "Reds", direction = 1, na.value = "grey90",
                       name = "Median\nduration\n(min)") +
  labs(x = "To category", y = "From category",
       title = "Median duration before switching to each category") +
  theme_minimal() +
  theme(axis.text = element_text(size = 11),
        axis.title = element_text(size = 11))
```

---

## Category-level process map

```{r cat-eventlog}
df_bupar_cat <- df_log_raw |>
  mutate(
    activity_instance = row_number(),
    lifecycle         = "complete",
    resource          = "all"
  )

el_cat <- eventlog(
  df_bupar_cat,
  case_id              = "case_id",
  activity_id          = "app_category",
  activity_instance_id = "activity_instance",
  lifecycle_id         = "lifecycle",
  timestamp            = "start_time",
  resource_id          = "resource"
)
```

### Start and end activities

Which category typically opens and closes a person-day?

```{r start-end-cat}
bind_rows(
  el_cat |> start_activities(level = "activity") |> mutate(position = "start"),
  el_cat |> end_activities(level = "activity")   |> mutate(position = "end")
) |>
  select(app_category, position, absolute, relative) |>
  arrange(position, desc(absolute)) |>
  knitr::kable(
    digits = 3,
    caption = "Which category begins and ends a person-day (relative = share of cases)"
  )
```

```{r cat-process-map, fig.width=8, fig.height=6}
tryCatch(
  el_cat |> process_map(type = frequency("relative"), render = TRUE),
  error = function(e) message("process_map cat: ", conditionMessage(e))
)
```

---

## Summary of key outputs

| Output | What it shows |
|--------|--------------|
| Activity frequency | Which apps appear most in the event log |
| Case length distribution | How many app uses per person-day |
| Dotted chart | All events across all cases at a glance |
| Hour / weekday distributions | Temporal rhythms of app use |
| Transition duration heatmap | How long in each category before switching |
| Start / end activities | Which category opens and closes a person-day |
| Category process map | Interpretable process model for the paper |
