---
title: "Process Mining of App Usage"
subtitle: "Method 5 — bupaR · ICA 2026 Hackathon"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    self-contained: true
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(bupaR)
library(processmapR)
if (requireNamespace("heuristicsmineR", quietly = TRUE)) library(heuristicsmineR)
library(scales)
library(patchwork)
source("utils.R")
```

## Conceptual framing

Process mining treats app usage as a **business process**: each person-day
is a *case*, each app use is an *event*, and the sequence of events forms a
*trace*. The goal is to discover recurring process patterns — frequent routes
through the app landscape — and to visualise them as process models.

The reference study (Fan et al., 2026) used `pm4py` in Python plus ProM Lite.
We replicate the key analyses using the `bupaR` ecosystem in R:

| Reference (pm4py / ProM) | This file (bupaR) |
|--------------------------|-------------------|
| `format_dataframe()` | `eventlog()` |
| `filter_variants_top_k()` | `filter_top_variants()` |
| `discover_dfg()` | `process_map(frequency())` |
| `discover_heuristics_net()` | `heuristics_map()` |
| `discover_bpmn_inductive()` | `process_map()` on filtered log |
| Event distribution plots | `dotted_chart()`, manual ggplot |

**Unit (case):** person-day — all app uses for one person on one calendar
day. This keeps traces short enough for process models to be readable.

**Activity:** `app_name` (fine-grained) or `app_category` (coarser).

**Activity:** direct app usage episodes — Pixel Launcher, Xperia Home, and
any other home-screen artefacts are excluded (see `utils.R`).

---

## Data

```{r load-data}
df_raw <- read_csv(
  "../d_apps_two_participants.csv",
  col_types = cols(
    participant_code = col_character(),
    start_time       = col_datetime(),
    end_time         = col_datetime(),
    duration         = col_double(),
    phone_session_id = col_double(),
    app_session_id   = col_double(),
    package_name     = col_character(),
    app_name         = col_character(),
    category         = col_character(),
    .default         = col_skip()
  )
) |>
  rename(app_category = category, Duration = duration) |>
  mutate(app_category = translate_category(app_category)) |>
  arrange(participant_code, start_time)
```

---

## Data preparation

### Merge consecutive same-app episodes

```{r merge}
df_merged <- df_raw |>
  filter(Duration > 0) |>
  group_by(participant_code) |>
  group_modify(~ merge_consecutive(.x, gap_threshold)) |>
  ungroup() |>
  filter(!app_name %in% noise_apps)  # remove launcher/home-screen artefacts (see utils.R)
```

### Build event log format

Cases are person-days. Each app-use episode is one event.

```{r build-events}
df_events <- df_merged |>
  mutate(date = as.Date(start_time)) |>
  group_by(participant_code, date) |>
  mutate(
    case_id  = paste0(participant_code, "_", date),
    next_app = lead(app_name)
  ) |>
  ungroup()

df_log_raw <- df_events |>
  arrange(case_id, start_time) |>
  select(case_id, participant_code, app_name, app_category,
         start_time, end_time, Duration)

# Filter case lengths: 3 < events < 1000 (upper bound calibrated to this dataset)
case_lengths <- df_log_raw |> count(case_id, name = "n_events")

cat("Case length distribution (app events per person-day):\n")
summary(case_lengths$n_events) |> print()

valid_cases <- case_lengths |> filter(n_events > 3, n_events < 1000) |> pull(case_id)

df_log_raw <- df_log_raw |> filter(case_id %in% valid_cases)

cat(sprintf(
  "Event log: %d cases, %d events, %d unique activities\n",
  n_distinct(df_log_raw$case_id),
  nrow(df_log_raw),
  n_distinct(df_log_raw$app_name)
))
```

### Create bupaR event log

```{r create-eventlog}
# bupaR requires: case_id, activity_id, activity_instance_id,
#                 lifecycle (start/complete), timestamp, resource
df_bupar <- df_log_raw |>
  mutate(
    # Each row is a complete activity instance
    activity_instance = row_number(),
    lifecycle         = "complete",
    resource          = app_category
  )

el <- eventlog(
  df_bupar,
  case_id              = "case_id",
  activity_id          = "app_name",
  activity_instance_id = "activity_instance",
  lifecycle_id         = "lifecycle",
  timestamp            = "start_time",
  resource_id          = "resource"
)

summary(el)
```

---

## Exploratory process analysis

```{r activity-frequency}
# Most frequent activities across all cases
el |>
  activity_frequency("activity") |>
  arrange(desc(absolute)) |>
  head(15) |>
  knitr::kable(caption = "Top 15 most frequent activities")
```

```{r case-length-dist, fig.width=7, fig.height=3}
df_log_raw |>
  count(case_id, name = "trace_length") |>
  ggplot(aes(trace_length)) +
  geom_histogram(bins = 30, fill = "#4E79A7", colour = "white") +
  labs(x = "Events per case (person-day)", y = "Count",
       title = "Case length distribution") +
  theme_minimal()
```

```{r dotted-chart, fig.width=10, fig.height=6}
# Dotted chart: one row per case, dots = events, x = time within case
# Coloured by app_category (resource field) — 5 colours instead of 116
tryCatch(
  print(el |>
    dotted_chart(x = "relative", color = "resource") +
    scale_color_manual(values = cat_pal) +
    labs(title = "Dotted chart: app category events within each person-day case",
         colour = "Category")),
  error = function(e) message("dotted_chart: ", conditionMessage(e))
)
```

---

## Process maps (Directly Follows Graph)

```{r filter-top-activities}
# With 100+ activities the full DFG is an unreadable hairball.
# Keep only the top-N most frequent apps.
top_n_activities <- 20  # DECISION POINT

top_acts <- df_bupar |>
  count(app_name, sort = TRUE) |>
  slice_head(n = top_n_activities) |>
  pull(app_name)

df_bupar_small <- df_bupar |>
  filter(app_name %in% top_acts) |>
  mutate(activity_instance = row_number())

el_small <- eventlog(
  df_bupar_small,
  case_id              = "case_id",
  activity_id          = "app_name",
  activity_instance_id = "activity_instance",
  lifecycle_id         = "lifecycle",
  timestamp            = "start_time",
  resource_id          = "resource"
)

cat(sprintf(
  "Reduced log: top %d activities, %d events retained of %d total\n",
  top_n_activities, nrow(df_bupar_small), nrow(df_bupar)
))
```

### Session-level process map (primary)

Median trace length at the phone-session level is ~15 events — far more
appropriate for process mining than person-day traces (~276 events).
This session-level map is the primary analysis; the person-day map below
is exploratory.

**Caveat:** `phone_session_id` is documented as non-monotonic with unclear
reset logic. Session boundaries here are taken at face value; validate against
time-gap heuristics (e.g., gap > 30 min = new session) before publishing.

```{r session-eventlog}
# Build session-level log from df_merged (which retains phone_session_id).
# df_log_raw drops phone_session_id in its select(), so we go back to the source.
df_bupar_sess <- df_merged |>
  mutate(
    case_id_sess      = paste0(participant_code, "_", phone_session_id),
    activity_instance = row_number(),
    lifecycle         = "complete",
    resource          = app_category
  )

el_sess <- eventlog(
  df_bupar_sess,
  case_id              = "case_id_sess",
  activity_id          = "app_name",
  activity_instance_id = "activity_instance",
  lifecycle_id         = "lifecycle",
  timestamp            = "start_time",
  resource_id          = "resource"
)

# Apply the same top-N activity filter as the person-day log
df_bupar_sess_small <- df_bupar_sess |>
  filter(app_name %in% top_acts) |>
  mutate(activity_instance = row_number())

el_sess_small <- eventlog(
  df_bupar_sess_small,
  case_id              = "case_id_sess",
  activity_id          = "app_name",
  activity_instance_id = "activity_instance",
  lifecycle_id         = "lifecycle",
  timestamp            = "start_time",
  resource_id          = "resource"
)

cat(sprintf(
  "Session log: %d sessions, %d events (top-%d activities)\n",
  bupaR::n_cases(el_sess_small), nrow(df_bupar_sess_small), top_n_activities
))
```

```{r session-process-map, fig.width=10, fig.height=7}
tryCatch(
  el_sess_small |> process_map(type = frequency("relative"), render = TRUE),
  error = function(e) message("session process_map: ", conditionMessage(e))
)
```

### Person-day process map (exploratory)

```{r process-map-freq, fig.width=10, fig.height=7}
tryCatch(
  el_small |> process_map(type = frequency("relative"), render = TRUE),
  error = function(e) message("process_map: ", conditionMessage(e))
)
```

```{r process-map-perf, fig.width=10, fig.height=7}
# Performance-based: mean duration per activity (top-N apps only)
el_timed <- df_bupar_small |>
  filter(!is.na(Duration)) |>
  mutate(activity_instance = row_number()) |>
  eventlog(
    case_id = "case_id", activity_id = "app_name",
    activity_instance_id = "activity_instance",
    lifecycle_id = "lifecycle", timestamp = "start_time",
    resource_id = "resource"
  )

tryCatch(
  el_timed |> process_map(type = performance(mean, units = "secs"), render = TRUE),
  error = function(e) message("process_map perf: ", conditionMessage(e))
)
```

### Filter to top-k variants

```{r top-variants}
# With 116 activities every person-day is a unique trace.
# Filter by trace frequency on the reduced log where common sub-patterns can emerge.
el_top10 <- tryCatch(
  edeaR::filter_trace_frequency(el_small, percentage = 0.8),
  error = function(e) { message("filter_trace_frequency: ", conditionMessage(e)); el_small }
)

cat(sprintf(
  "Filtered log: %d of %d cases\n",
  bupaR::n_cases(el_top10), bupaR::n_cases(el_small)
))

tryCatch(
  print(edeaR::trace_coverage(el_top10, "case") |>
    head(10) |>
    knitr::kable(caption = "Most frequent trace variants (top-20 activities)")),
  error = function(e) message("trace_coverage: ", conditionMessage(e))
)
```

```{r top-variants-map, fig.width=10, fig.height=7}
tryCatch(
  el_top10 |> process_map(type = frequency("relative"), render = TRUE),
  error = function(e) message("process_map top10: ", conditionMessage(e))
)
```

---

## Heuristics miner

The heuristics miner is more noise-tolerant than the inductive miner and
produces a *heuristics net* showing the most reliable activity connections.

```{r heuristics-map, fig.width=10, fig.height=7}
if (requireNamespace("heuristicsmineR", quietly = TRUE)) {
  el_small |> heuristicsmineR::heuristics_map(dependency_threshold = 0.5)
} else {
  message("heuristicsmineR not available — skipping heuristics map")
}
```

---

## Event distribution analysis

```{r hour-distribution, fig.width=8, fig.height=4}
df_log_raw |>
  mutate(hour = as.integer(format(start_time, "%H"))) |>
  count(hour, app_category) |>
  group_by(hour) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(hour, pct, fill = app_category)) +
  geom_col() +
  scale_x_continuous(breaks = seq(0, 23, 2)) +
  scale_fill_manual(values = cat_pal) +
  labs(x = "Hour of day", y = "Proportion of events",
       title = "App category usage by hour of day", fill = "Category") +
  theme_minimal()
```

```{r weekday-distribution, fig.width=8, fig.height=4}
df_log_raw |>
  mutate(weekday = weekdays(as.Date(start_time)),
         weekday = factor(weekday, levels = c(
           "Monday","Tuesday","Wednesday","Thursday","Friday","Saturday","Sunday"))) |>
  count(weekday, app_category) |>
  group_by(weekday) |>
  mutate(pct = n / sum(n)) |>
  ungroup() |>
  ggplot(aes(weekday, pct, fill = app_category)) +
  geom_col() +
  scale_fill_manual(values = cat_pal) +
  labs(x = NULL, y = "Proportion", fill = "Category",
       title = "App category usage by day of week") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

---

## Category transition duration heatmap

How long (median minutes) do people spend in each category before switching
to each other category? Mirrors the platform transition heatmap in the
reference (their `duration_time.csv`).

```{r transition-heatmap, fig.width=6, fig.height=5}
cat_pal_rocket <- c(
  "social network" = "#F28E2B", "communication" = "#4E79A7",
  "tools" = "#59A14F", "entertainment" = "#B07AA1"
)
core_cats <- names(cat_pal_rocket)

df_transitions <- df_merged |>
  filter(app_category %in% core_cats) |>
  arrange(participant_code, start_time) |>
  group_by(participant_code) |>
  mutate(
    next_cat = lead(app_category),
    dur_min  = Duration / 60
  ) |>
  ungroup() |>
  filter(!is.na(next_cat), app_category != next_cat,
         next_cat %in% core_cats)

heatmap_data <- df_transitions |>
  group_by(from = app_category, to = next_cat) |>
  summarise(median_dur_min = median(dur_min), .groups = "drop") |>
  complete(from = core_cats, to = core_cats, fill = list(median_dur_min = NA))

heatmap_data |>
  ggplot(aes(to, from, fill = median_dur_min)) +
  geom_tile(colour = "white", linewidth = 1) +
  geom_text(aes(label = ifelse(is.na(median_dur_min), "—",
                               sprintf("%.1f", median_dur_min))),
            size = 4) +
  scale_fill_distiller(palette = "Reds", direction = 1, na.value = "grey90",
                       name = "Median\nduration\n(min)") +
  labs(x = "To category", y = "From category",
       title = "Median duration before switching to each category") +
  theme_minimal() +
  theme(axis.text = element_text(size = 11),
        axis.title = element_text(size = 11))
```

---

## Category-level process map

A coarser, more interpretable process map using `app_category` as the
activity — cleaner for a paper figure than the app-level map.

```{r cat-eventlog}
df_bupar_cat <- df_log_raw |>
  mutate(
    activity_instance = row_number(),
    lifecycle         = "complete",
    resource          = "all"
  )

el_cat <- eventlog(
  df_bupar_cat,
  case_id              = "case_id",
  activity_id          = "app_category",
  activity_instance_id = "activity_instance",
  lifecycle_id         = "lifecycle",
  timestamp            = "start_time",
  resource_id          = "resource"
)
```

```{r cat-process-map, fig.width=8, fig.height=6}
tryCatch(
  el_cat |> process_map(type = frequency("relative"), render = TRUE),
  error = function(e) message("process_map cat: ", conditionMessage(e))
)
```

---

## Summary of key outputs

| Output | What it shows |
|--------|--------------|
| Activity frequency | Which apps appear most in the event log |
| Case length distribution | How many app uses per person-day |
| Dotted chart | All events across all cases at a glance |
| Frequency process map | Most common app-to-app flows (= DFG) |
| Performance process map | Mean duration spent per app |
| Top-10 variant map | Most frequent full trace patterns |
| Heuristics net | Noise-tolerant process model |
| Hour / weekday distributions | Temporal rhythms of app use |
| Transition duration heatmap | How long in each category before switching |
| Category process map | Coarse-grained process model for the paper |
