---
title: "Language-Based Models of App Usage"
subtitle: "Method 6 — Activity2Vec / Doc2Vec · ICA 2026 Hackathon"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    self-contained: true
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(word2vec)
library(doc2vec)
library(Rtsne)
library(umap)
library(cluster)
library(scales)
library(patchwork)
library(ggrepel)
library(RColorBrewer)
```

## Conceptual framing

The language-model approach treats app sequences as *text*: each app use is
a "word", each person's full history is a "document". Trained on the corpus
of all sequences, a Word2Vec model learns a **dense embedding** for each app
such that apps that appear in similar sequential contexts end up close in
embedding space.

The reference study (Fan et al., 2026) used `gensim`'s Word2Vec with rich
tokens combining platform + interaction content + activity type. Because our
data contains only `app_name` and `app_category` (no content), we enrich
the token with **time-of-day context** — "Instagram\_morning" vs.
"Instagram\_evening" — to give the model more signal.

Two models:

1. **Activity2Vec** (Word2Vec on tokens): learns which apps are
   contextually similar — i.e., tend to appear in similar sequential
   neighbourhoods across users.
2. **Person2Vec** (Doc2Vec on full sequences): produces a single embedding
   per person, enabling person-level clustering and comparison.

---

## Data

```{r load-data}
# --- Real data (uncomment when available) ---
# app_categories <- read_csv("data/raw/app_categories.csv")
# df_raw <- read_csv(
#   "data/raw/uni_mainz_app_session.csv",
#   col_types = cols(
#     participant_code = col_character(),
#     start_time       = col_datetime(),
#     Duration         = col_double(),
#     package_name     = col_character(),
#     app_name         = col_character(),
#     .default         = col_skip()
#   )
# ) |>
#   left_join(app_categories, by = "package_name") |>
#   arrange(participant_code, start_time)

# --- SYNTHETIC SAMPLE ---
set.seed(42)
pkg_to_app <- c(
  "com.instagram.android" = "Instagram", "com.whatsapp" = "WhatsApp",
  "com.zhiliaoapp.musically" = "TikTok", "com.google.android.youtube" = "YouTube",
  "com.android.chrome" = "Chrome", "com.google.android.apps.maps" = "Maps",
  "com.google.android.calculator" = "Rechner", "com.google.android.gm" = "Gmail",
  "com.facebook.orca" = "Messenger", "com.spotify.music" = "Spotify"
)
pkg_to_cat <- c(
  "com.instagram.android" = "social network", "com.whatsapp" = "communication",
  "com.zhiliaoapp.musically" = "social network", "com.google.android.youtube" = "social network",
  "com.android.chrome" = "tools", "com.google.android.apps.maps" = "tools",
  "com.google.android.calculator" = "tools", "com.google.android.gm" = "communication",
  "com.facebook.orca" = "communication", "com.spotify.music" = "entertainment"
)
pkg_names <- names(pkg_to_app); pkg_probs <- c(3, 3, 2, 2, 1, 1, .5, 1, 1, 1)

df_raw <- map2_dfr(
  paste0("P", sprintf("%03d", 1:200)),
  sample(100:600, 200, replace = TRUE),
  function(pid, n) {
    starts <- as.POSIXct("2024-04-01 08:00:00") +
      cumsum(c(0, sample(30:3600, n - 1, replace = TRUE)))
    pkgs <- sample(pkg_names, n, replace = TRUE, prob = pkg_probs)
    tibble(
      participant_code = pid,
      package_name = pkgs, app_name = pkg_to_app[pkgs],
      app_category = pkg_to_cat[pkgs],
      start_time = starts,
      Duration = sample(10:900, n, replace = TRUE) + runif(n)
    )
  }
) |> arrange(participant_code, start_time)
# --- end synthetic sample ---
```

---

## Data preparation

### Merge consecutive same-app episodes

```{r merge}
gap_threshold <- 10 * 60

merge_consecutive <- function(d, gap_sec) {
  d <- arrange(d, start_time); out <- list(); cur <- d[1, ]
  for (i in seq_len(nrow(d))[-1]) {
    row <- d[i, ]
    gap <- as.numeric(row$start_time - (cur$start_time + cur$Duration), units = "secs")
    if (row$app_name == cur$app_name && !is.na(gap) && gap <= gap_sec) {
      cur$Duration <- cur$Duration + row$Duration
    } else { out <- c(out, list(cur)); cur <- row }
  }
  bind_rows(c(out, list(cur)))
}

df_merged <- df_raw |>
  filter(Duration > 0) |>
  group_by(participant_code) |>
  group_modify(~ merge_consecutive(.x, gap_threshold)) |>
  ungroup()
```

### Create enriched tokens

Because our app vocabulary is small (~10–50 apps), raw app names give
Word2Vec little contextual variation. We enrich each token with a time-of-day
label, producing tokens like `Instagram_evening` or `YouTube_afternoon`.

```{r tokens}
df_tokens <- df_merged |>
  mutate(
    hour = as.integer(format(start_time, "%H")),
    time_of_day = case_when(
      hour >= 6  & hour < 12 ~ "morning",
      hour >= 12 & hour < 17 ~ "afternoon",
      hour >= 17 & hour < 22 ~ "evening",
      TRUE                   ~ "night"
    ),
    # Base token: just the app name
    token_app  = str_replace_all(app_name, " ", "_"),
    # Enriched token: app + time of day (more variation for Word2Vec)
    token_rich = paste0(token_app, "_", time_of_day)
  )

cat("Unique base tokens:    ", n_distinct(df_tokens$token_app),  "\n")
cat("Unique enriched tokens:", n_distinct(df_tokens$token_rich), "\n")
cat("Total events:          ", nrow(df_tokens), "\n")
```

### Build per-person sequences

```{r sequences}
# Base: one character vector of app tokens per person
seqs_app <- df_tokens |>
  arrange(participant_code, start_time) |>
  group_by(participant_code) |>
  summarise(tokens = list(token_app), .groups = "drop")

# Enriched: app + time-of-day
seqs_rich <- df_tokens |>
  arrange(participant_code, start_time) |>
  group_by(participant_code) |>
  summarise(tokens = list(token_rich), .groups = "drop")

# Word2Vec needs sequences as a flat character vector (sentences = " "-separated)
corpus_app  <- sapply(seqs_app$tokens,  paste, collapse = " ")
corpus_rich <- sapply(seqs_rich$tokens, paste, collapse = " ")

cat("Corpus size (persons):", length(corpus_app), "\n")
cat("Median sequence length:", median(lengths(seqs_app$tokens)), "tokens\n")
```

---

## Activity2Vec (Word2Vec)

### Train model

```{r train-w2v, cache=TRUE}
# DECISION POINT: dim, window, min_count, iter
# - dim: 50 sufficient for small vocabularies; use 100+ for large ones
# - window: 5 matches the reference; smaller = more local context
# - min_count: minimum token frequency to include; 2 is reasonable for small data
# - type: "skip-gram" (sg=1 in reference) better for rare tokens

w2v_path <- "results/tables/w2v_model.bin"

if (file.exists(w2v_path)) {
  w2v <- read.word2vec(w2v_path)
} else {
  set.seed(42)
  w2v <- word2vec(
    x         = corpus_rich,
    type      = "skip-gram",
    dim       = 50,
    window    = 5,
    min_count = 2,
    iter      = 20,
    threads   = parallel::detectCores() - 1
  )
  write.word2vec(w2v, w2v_path)
}

vocab <- summary(w2v, type = "vocabulary")
cat("Vocabulary size:", length(vocab), "\n")
```

### Explore: most similar tokens

```{r similar-tokens}
# For each app, find its nearest neighbours in embedding space
# Expected: same-app-different-time variants should cluster together
# Cross-app: apps used in similar behavioural contexts should be close

find_similar <- function(token, model, n = 5) {
  tryCatch(
    predict(model, token, type = "nearest", top_n = n) |>
      as_tibble() |>
      mutate(query = token) |>
      select(query, term2, similarity),
    error = function(e) tibble(query = token, term2 = NA, similarity = NA)
  )
}

# Query a few representative tokens
query_tokens <- vocab[seq(1, min(6, length(vocab)))]
map_dfr(query_tokens, find_similar, model = w2v) |>
  knitr::kable(digits = 3, caption = "Nearest neighbours in embedding space")
```

### Extract and visualise embeddings

```{r extract-embeddings}
emb_mat <- as.matrix(w2v)

# Parse token back into app name and time-of-day
emb_df <- as_tibble(emb_mat, rownames = "token") |>
  mutate(
    parts       = str_split(token, "_(?=morning|afternoon|evening|night)"),
    app_name    = map_chr(parts, 1),
    time_of_day = map_chr(parts, \(p) if (length(p) > 1) p[2] else "unknown"),
    app_category = pkg_to_cat[
      names(pkg_to_app)[match(app_name, pkg_to_app)]
    ]
  ) |>
  select(-parts)
```

```{r tsne-embeddings, fig.width=9, fig.height=7}
# t-SNE dimensionality reduction (mirrors the reference visualisation)
# Only run if enough tokens for t-SNE (needs > perplexity * 3)
if (nrow(emb_mat) > 15) {
  set.seed(42)
  tsne_out <- Rtsne(
    emb_mat,
    dims       = 2,
    perplexity = min(5, floor((nrow(emb_mat) - 1) / 3)),
    verbose    = FALSE,
    max_iter   = 1000
  )

  cat_pal <- c(
    "social network" = "#F28E2B", "communication" = "#4E79A7",
    "tools" = "#59A14F", "entertainment" = "#B07AA1", "other" = "#BAB0AC"
  )
  tod_shapes <- c(morning = 16, afternoon = 17, evening = 15, night = 18, unknown = 3)

  emb_df |>
    mutate(tsne1 = tsne_out$Y[, 1], tsne2 = tsne_out$Y[, 2]) |>
    ggplot(aes(tsne1, tsne2, colour = app_category, shape = time_of_day,
               label = token)) +
    geom_point(size = 3, alpha = 0.85) +
    ggrepel::geom_text_repel(size = 2.8, max.overlaps = 12) +
    scale_colour_manual(values = cat_pal, na.value = "#BAB0AC", name = "Category") +
    scale_shape_manual(values = tod_shapes, name = "Time of day") +
    labs(title = "Activity2Vec embedding space (t-SNE)",
         subtitle = "Tokens closer together appear in similar sequential contexts",
         x = "t-SNE 1", y = "t-SNE 2") +
    theme_minimal()
} else {
  cat("Too few tokens for t-SNE — use PCA instead or expand vocabulary.\n")
  # PCA fallback
  pca_out <- prcomp(emb_mat, scale. = TRUE)
  emb_df |>
    mutate(pc1 = pca_out$x[, 1], pc2 = pca_out$x[, 2]) |>
    ggplot(aes(pc1, pc2, colour = app_category, label = token)) +
    geom_point(size = 3) +
    ggrepel::geom_text_repel(size = 3) +
    scale_colour_manual(values = cat_pal, na.value = "#BAB0AC") +
    labs(title = "Activity2Vec embedding space (PCA)", x = "PC1", y = "PC2") +
    theme_minimal()
}
```

---

## Person2Vec (Doc2Vec)

Doc2Vec extends Word2Vec to produce a single embedding per person
(treating each person's full sequence as a "document"). This enables
**person-level** analysis: who is similar to whom, and do persons cluster
into meaningful behavioural types?

### Train model

```{r train-d2v, cache=TRUE}
d2v_path <- "results/tables/d2v_model.bin"

if (file.exists(d2v_path)) {
  d2v <- read.paragraph2vec(d2v_path)
} else {
  # doc2vec expects a data frame with "doc_id" and "text" columns
  d2v_input <- tibble(
    doc_id = seqs_rich$participant_code,
    text   = corpus_rich
  )

  set.seed(42)
  d2v <- paragraph2vec(
    x         = d2v_input,
    type      = "PV-DM",    # DM = distributed memory (default, context-aware)
    dim       = 50,
    window    = 5,
    min_count = 2,
    iter      = 20,
    threads   = parallel::detectCores() - 1
  )
  write.paragraph2vec(d2v, d2v_path)
}
```

### Person embeddings: UMAP visualisation

```{r person-embeddings}
# Extract document (person) embeddings
person_emb <- as.matrix(d2v, which = "docs")
rownames(person_emb) <- seqs_rich$participant_code

cat("Person embedding matrix:", nrow(person_emb), "persons ×",
    ncol(person_emb), "dimensions\n")
```

```{r umap-persons, fig.width=9, fig.height=7}
set.seed(42)
umap_out <- umap(person_emb, n_neighbors = min(15, nrow(person_emb) - 1))

# Annotate with person-level usage stats (dominant category, sequence length)
person_stats <- df_tokens |>
  group_by(participant_code) |>
  summarise(
    seq_len       = n(),
    dominant_cat  = names(which.max(table(app_category))),
    pct_social    = mean(app_category == "social network"),
    pct_comm      = mean(app_category == "communication"),
    .groups = "drop"
  )

umap_df <- tibble(
  participant_code = rownames(person_emb),
  umap1 = umap_out$layout[, 1],
  umap2 = umap_out$layout[, 2]
) |>
  left_join(person_stats, by = "participant_code")

cat_pal <- c(
  "social network" = "#F28E2B", "communication" = "#4E79A7",
  "tools" = "#59A14F", "entertainment" = "#B07AA1", "other" = "#BAB0AC"
)

umap_df |>
  ggplot(aes(umap1, umap2, colour = dominant_cat, size = seq_len)) +
  geom_point(alpha = 0.75) +
  scale_colour_manual(values = cat_pal, name = "Dominant\ncategory") +
  scale_size(range = c(1.5, 6), name = "Sequence\nlength") +
  labs(title = "Person2Vec embedding space (UMAP)",
       subtitle = "Each point is one participant; proximity = similar usage patterns",
       x = "UMAP 1", y = "UMAP 2") +
  theme_minimal()
```

### Cluster persons in embedding space

```{r person-clusters, fig.width=9, fig.height=7}
# DECISION POINT: choose k via silhouette (same approach as Method 1)
k_range <- 2:6
sil_widths <- sapply(k_range, function(k) {
  cl  <- kmeans(person_emb, centers = k, nstart = 25, iter.max = 100)
  sil <- silhouette(cl$cluster, dist(person_emb))
  mean(sil[, 3])
})

tibble(k = k_range, avg_silhouette = sil_widths) |>
  ggplot(aes(k, avg_silhouette)) +
  geom_line() + geom_point(size = 2) +
  labs(title = "Person cluster quality by k (k-means on Doc2Vec embeddings)",
       x = "k", y = "Average silhouette width") +
  theme_minimal()
```

```{r cluster-umap, fig.width=9, fig.height=7}
k_chosen <- k_range[which.max(sil_widths)]
cat("Chosen k:", k_chosen, "\n")

set.seed(42)
km <- kmeans(person_emb, centers = k_chosen, nstart = 25)

umap_df |>
  mutate(cluster = factor(km$cluster)) |>
  ggplot(aes(umap1, umap2, colour = cluster, size = seq_len)) +
  geom_point(alpha = 0.8) +
  scale_colour_brewer(palette = "Set2", name = "Cluster") +
  scale_size(range = c(1.5, 6), name = "Sequence\nlength") +
  labs(title = paste0("Person clusters in Doc2Vec space (k = ", k_chosen, ")"),
       x = "UMAP 1", y = "UMAP 2") +
  theme_minimal()
```

```{r cluster-profiles}
# What characterises each cluster?
tibble(
  participant_code = rownames(person_emb),
  cluster = factor(km$cluster)
) |>
  left_join(person_stats, by = "participant_code") |>
  group_by(cluster) |>
  summarise(
    n          = n(),
    med_seq_len = median(seq_len),
    pct_social  = round(mean(pct_social) * 100, 1),
    pct_comm    = round(mean(pct_comm) * 100, 1),
    dominant    = names(which.max(table(dominant_cat))),
    .groups = "drop"
  ) |>
  knitr::kable(caption = "Cluster profiles")
```

---

## User trajectory visualisation

Plot a single person's sequence as a trajectory through the embedding
space — each point is one app use, arrows show the temporal order.
Mirrors the quiver-plot visualisation in the reference.

```{r trajectory, fig.width=9, fig.height=7}
# Pick one person for the trajectory (choose a person with a long sequence)
trajectory_person <- person_stats |>
  slice_max(seq_len, n = 1) |>
  pull(participant_code)

# Get their token sequence and look up embeddings
traj_tokens <- df_tokens |>
  filter(participant_code == trajectory_person) |>
  arrange(start_time) |>
  pull(token_rich)

# Only keep tokens that are in the Word2Vec vocabulary
in_vocab <- traj_tokens %in% rownames(emb_mat)
traj_emb <- emb_mat[traj_tokens[in_vocab], , drop = FALSE]

if (nrow(traj_emb) >= 4) {
  # Project to 2D with PCA fitted on all embeddings
  pca_all  <- prcomp(emb_mat, scale. = TRUE)
  traj_2d  <- predict(pca_all, traj_emb)[, 1:2]
  all_2d   <- pca_all$x[, 1:2]

  # Background: all token embeddings
  bg_df <- as_tibble(all_2d, rownames = "token") |>
    mutate(
      app_name = map_chr(str_split(token, "_(?=morning|afternoon|evening|night)"), 1),
      app_category = pkg_to_cat[names(pkg_to_app)[match(app_name, pkg_to_app)]]
    )

  # Trajectory data frame: one row per step
  traj_df <- as_tibble(traj_2d) |>
    rename(x = PC1, y = PC2) |>
    mutate(
      t        = row_number(),
      token    = traj_tokens[in_vocab],
      app_name = map_chr(str_split(token, "_(?=morning|afternoon|evening|night)"), 1)
    )

  ggplot() +
    # Background cloud of all token embeddings
    geom_point(data = bg_df, aes(PC1, PC2, colour = app_category),
               alpha = 0.2, size = 2) +
    # Trajectory arrows
    geom_segment(
      data = traj_df |> mutate(xend = lead(x), yend = lead(y)) |> filter(!is.na(xend)),
      aes(x = x, y = y, xend = xend, yend = yend, colour = factor(t %% 8)),
      arrow = arrow(length = unit(2.5, "mm"), type = "closed"),
      linewidth = 0.7, show.legend = FALSE
    ) +
    # Points
    geom_point(data = traj_df, aes(x, y), colour = "black", size = 1.5) +
    scale_colour_manual(
      values = c("social network" = "#F28E2B", "communication" = "#4E79A7",
                 "tools" = "#59A14F", "entertainment" = "#B07AA1",
                 "other" = "#BAB0AC",
                 # trajectory step colours (recycled, hidden from legend)
                 setNames(RColorBrewer::brewer.pal(8, "PuRd"), as.character(0:7))),
      name = "Category"
    ) +
    labs(
      title = paste0("Trajectory of participant ", trajectory_person,
                     " through activity embedding space"),
      subtitle = "Arrows show temporal order of app uses",
      x = "PC1", y = "PC2"
    ) +
    theme_minimal() +
    guides(colour = guide_legend(override.aes = list(size = 4, alpha = 1)))
} else {
  cat("Not enough vocabulary overlap for trajectory plot.\n")
}
```

---

## Summary of key outputs

| Output | What it shows |
|--------|--------------|
| Vocabulary & token counts | How rich the token space is with/without time enrichment |
| Nearest neighbours table | Which apps appear in similar sequential contexts |
| t-SNE embedding plot | Activity embedding space — categories should cluster |
| UMAP person plot | Person-level behavioural space, coloured by dominant category |
| Silhouette plot | How many person clusters are supported |
| Cluster UMAP | Person clusters in embedding space |
| Cluster profiles | Usage characteristics of each behavioural cluster |
| Trajectory plot | One person's path through the embedding space over time |
