---
title: "Language-Based Models of App Usage"
subtitle: "Method 6 — Activity2Vec / Doc2Vec · ICA 2026 Hackathon"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    self-contained: true
execute:
  warning: false
  message: false
---

```{r setup}
library(tidyverse)
library(word2vec)
library(doc2vec)
library(Rtsne)
library(umap)
library(cluster)
library(scales)
library(patchwork)
library(ggrepel)
library(RColorBrewer)
source("utils.R")
```

## Conceptual framing

The language-model approach treats app sequences as *text*: each app use is
a "word", each person's full history is a "document". Trained on the corpus
of all sequences, a Word2Vec model learns a **dense embedding** for each app
such that apps that appear in similar sequential contexts end up close in
embedding space.

The reference study (Fan et al., 2026) used `gensim`'s Word2Vec with rich
tokens combining platform + interaction content + activity type. Because our
data contains only `app_name` and `app_category` (no content), we enrich
the token with **time-of-day context** — "Instagram\_morning" vs.
"Instagram\_evening" — to give the model more signal.

Two models:

1. **Activity2Vec** (Word2Vec on tokens): learns which apps are
   contextually similar — i.e., tend to appear in similar sequential
   neighbourhoods across users.
2. **Person2Vec** (Doc2Vec on full sequences): produces a single embedding
   per person, enabling person-level clustering and comparison.

---

## Data

```{r load-data}
df_raw <- read_csv(
  "../d_apps_two_participants.csv",
  col_types = cols(
    participant_code = col_character(),
    start_time       = col_datetime(),
    end_time         = col_datetime(),
    duration         = col_double(),
    phone_session_id = col_double(),
    app_session_id   = col_double(),
    package_name     = col_character(),
    app_name         = col_character(),
    category         = col_character(),
    .default         = col_skip()
  )
) |>
  rename(app_category = category, Duration = duration) |>
  mutate(app_category = translate_category(app_category)) |>
  arrange(participant_code, start_time)
```

---

## Data preparation

### Merge consecutive same-app episodes

```{r merge}
df_merged <- df_raw |>
  filter(Duration > 0) |>
  group_by(participant_code) |>
  group_modify(~ merge_consecutive(.x, gap_threshold)) |>
  ungroup() |>
  filter(!app_name %in% noise_apps)  # remove launcher/home-screen artefacts (see utils.R)
```

### Create enriched tokens

Because our app vocabulary is small (~10–50 apps), raw app names give
Word2Vec little contextual variation. We enrich each token with a time-of-day
label, producing tokens like `Instagram_evening` or `YouTube_afternoon`.

```{r tokens}
df_tokens <- df_merged |>
  mutate(
    hour = as.integer(format(start_time, "%H")),
    time_of_day = case_when(
      hour >= 6  & hour < 12 ~ "morning",    # 06–12
      hour >= 12 & hour < 18 ~ "afternoon",  # 12–18
      hour >= 18 & hour < 24 ~ "evening",    # 18–24
      TRUE                   ~ "night"       # 00–06
    ),
    # Base token: just the app name
    token_app  = str_replace_all(app_name, " ", "_"),
    # Enriched token: app + time of day (more variation for Word2Vec)
    token_rich = paste0(token_app, "_", time_of_day)
  )

cat("Unique base tokens:    ", n_distinct(df_tokens$token_app),  "\n")
cat("Unique enriched tokens:", n_distinct(df_tokens$token_rich), "\n")
cat("Total events:          ", nrow(df_tokens), "\n")
```

### Build per-person sequences

```{r sequences}
# Base: one character vector of app tokens per person (for Word2Vec corpus)
seqs_app <- df_tokens |>
  arrange(participant_code, start_time) |>
  group_by(participant_code) |>
  summarise(tokens = list(token_app), .groups = "drop")

# Enriched: app + time-of-day (for Word2Vec corpus)
seqs_rich <- df_tokens |>
  arrange(participant_code, start_time) |>
  group_by(participant_code) |>
  summarise(tokens = list(token_rich), .groups = "drop")

# Word2Vec corpus: flat character vectors per person
corpus_app  <- sapply(seqs_app$tokens,  paste, collapse = " ")
corpus_rich <- sapply(seqs_rich$tokens, paste, collapse = " ")

# Per-day documents for Doc2Vec: one document per person-day, then average
# to get a person embedding. Very long per-person documents regress toward
# the corpus mean; per-day units give the model more varied training signal.
seqs_rich_day <- df_tokens |>
  arrange(participant_code, start_time) |>
  mutate(date = as.Date(start_time)) |>
  group_by(participant_code, date) |>
  summarise(
    doc_id = paste0(first(participant_code), "_", first(date)),
    text   = paste(token_rich, collapse = " "),
    .groups = "drop"
  )

cat("Corpus size (persons):", length(corpus_app), "\n")
cat("Median sequence length:", median(lengths(seqs_app$tokens)), "tokens\n")
cat("Per-day documents:", nrow(seqs_rich_day), "\n")
```

---

## Activity2Vec (Word2Vec)

### Train model

```{r train-w2v}
# DECISION POINT: dim, window, min_count, iter
# - dim: 50 sufficient for small vocabularies; use 100+ for large ones
# - window: 5 matches the reference; smaller = more local context
# - min_count: minimum token frequency to include; 2 is reasonable for small data
# - type: "skip-gram" (sg=1 in reference) better for rare tokens

w2v_path <- "results/tables/w2v_model.bin"

w2v <- NULL
if (file.exists(w2v_path)) {
  w2v <- read.word2vec(w2v_path)
  # Validate: cached vocabulary should be a subset of current tokens (allow extras from
  # min_count filtering, but a much larger vocab signals a stale model from other data).
  cached_vocab    <- summary(w2v, type = "vocabulary")
  expected_tokens <- unique(unlist(seqs_rich$tokens))
  if (length(cached_vocab) > length(expected_tokens) * 3) {
    message("Stale Word2Vec cache (vocabulary much larger than current data) — retraining.")
    w2v <- NULL
  }
}
if (is.null(w2v)) {
  set.seed(42)
  w2v <- word2vec(
    x         = corpus_rich,
    type      = "skip-gram",
    dim       = 50,
    window    = 5,
    min_count = 2,
    iter      = 20,
    threads   = parallel::detectCores() - 1
  )
  write.word2vec(w2v, w2v_path)
}

vocab <- summary(w2v, type = "vocabulary")
cat("Vocabulary size:", length(vocab), "\n")
```

### Explore: most similar tokens

```{r similar-tokens}
# For each app, find its nearest neighbours in embedding space
# Expected: same-app-different-time variants should cluster together
# Cross-app: apps used in similar behavioural contexts should be close

find_similar <- function(token, model, n = 5) {
  tryCatch(
    predict(model, token, type = "nearest", top_n = n) |>
      as_tibble() |>
      mutate(query = token) |>
      select(query, term2, similarity),
    error = function(e) tibble(query = token, term2 = NA, similarity = NA)
  )
}

# Query a few representative tokens
query_tokens <- vocab[seq(1, min(6, length(vocab)))]
map_dfr(query_tokens, find_similar, model = w2v) |>
  knitr::kable(digits = 3, caption = "Nearest neighbours in embedding space")
```

### Extract and visualise embeddings

```{r extract-embeddings}
emb_mat <- as.matrix(w2v)

# Parse token back into app name and time-of-day
app_cat_lookup <- df_merged |> distinct(app_name, app_category) |> deframe()

emb_df <- as_tibble(emb_mat, rownames = "token") |>
  mutate(
    parts       = str_split(token, "_(?=morning|afternoon|evening|night)"),
    app_name    = map_chr(parts, 1),
    time_of_day = map_chr(parts, \(p) if (length(p) > 1) p[2] else "unknown"),
    app_category = app_cat_lookup[app_name]
  ) |>
  select(-parts)
```

```{r tsne-embeddings, fig.width=9, fig.height=7}
# t-SNE dimensionality reduction (mirrors the reference visualisation)
# Only run if enough tokens for t-SNE (needs > perplexity * 3)
if (nrow(emb_mat) > 15) {
  set.seed(42)
  tsne_out <- Rtsne(
    emb_mat,
    dims       = 2,
    perplexity = min(30, floor((nrow(emb_mat) - 1) / 3)),
    verbose    = FALSE,
    max_iter   = 1000
  )

  cat_pal <- c(
    "social network" = "#F28E2B", "communication" = "#4E79A7",
    "tools" = "#59A14F", "entertainment" = "#B07AA1", "other" = "#BAB0AC"
  )
  tod_shapes <- c(morning = 16, afternoon = 17, evening = 15, night = 18, unknown = 3)

  emb_df |>
    mutate(tsne1 = tsne_out$Y[, 1], tsne2 = tsne_out$Y[, 2]) |>
    ggplot(aes(tsne1, tsne2, colour = app_category, shape = time_of_day,
               label = token)) +
    geom_point(size = 3, alpha = 0.85) +
    ggrepel::geom_text_repel(size = 2.8, max.overlaps = 12) +
    scale_colour_manual(values = cat_pal, na.value = "#BAB0AC", name = "Category") +
    scale_shape_manual(values = tod_shapes, name = "Time of day") +
    labs(title = "Activity2Vec embedding space (t-SNE)",
         subtitle = "Tokens closer together appear in similar sequential contexts",
         x = "t-SNE 1", y = "t-SNE 2") +
    theme_minimal()
} else {
  cat("Too few tokens for t-SNE — use PCA instead or expand vocabulary.\n")
  # PCA fallback
  pca_out <- prcomp(emb_mat, scale. = TRUE)
  emb_df |>
    mutate(pc1 = pca_out$x[, 1], pc2 = pca_out$x[, 2]) |>
    ggplot(aes(pc1, pc2, colour = app_category, label = token)) +
    geom_point(size = 3) +
    ggrepel::geom_text_repel(size = 3) +
    scale_colour_manual(values = cat_pal, na.value = "#BAB0AC") +
    labs(title = "Activity2Vec embedding space (PCA)", x = "PC1", y = "PC2") +
    theme_minimal()
}
```

---

## Person2Vec (Doc2Vec)

Doc2Vec extends Word2Vec to produce a single embedding per person
(treating each person's full sequence as a "document"). This enables
**person-level** analysis: who is similar to whom, and do persons cluster
into meaningful behavioural types?

### Train model

```{r train-d2v}
d2v_path <- "results/tables/d2v_model.bin"

d2v <- NULL
if (file.exists(d2v_path)) {
  d2v <- read.paragraph2vec(d2v_path)
  # Validate: doc IDs must match current per-day document IDs.
  cached_doc_ids <- rownames(as.matrix(d2v, which = "docs"))
  if (!setequal(cached_doc_ids, seqs_rich_day$doc_id)) {
    message("Stale Doc2Vec cache (doc IDs don't match current per-day documents) — retraining.")
    d2v <- NULL
  }
}
if (is.null(d2v)) {
  d2v_input <- seqs_rich_day |> select(doc_id, text)

  set.seed(42)
  d2v <- paragraph2vec(
    x         = d2v_input,
    type      = "PV-DM",    # DM = distributed memory (default, context-aware)
    dim       = 50,
    window    = 5,
    min_count = 2,
    iter      = 20,
    threads   = parallel::detectCores() - 1
  )
  write.paragraph2vec(d2v, d2v_path)
}
```

### Person embeddings: UMAP visualisation

```{r person-embeddings}
# Average per-day embeddings to get person-level embeddings.
# Splitting into days gives more varied training signal; averaging recovers
# the person-level representation for downstream clustering/UMAP.
person_emb <- as.matrix(d2v, which = "docs") |>
  as_tibble(rownames = "doc_id") |>
  mutate(
    date             = str_extract(doc_id, "\\d{4}-\\d{2}-\\d{2}$"),
    participant_code = str_remove(doc_id, "_\\d{4}-\\d{2}-\\d{2}$")
  ) |>
  group_by(participant_code) |>
  summarise(across(where(is.numeric), mean), .groups = "drop") |>
  column_to_rownames("participant_code") |>
  as.matrix()

cat("Person embedding matrix:", nrow(person_emb), "persons ×",
    ncol(person_emb), "dimensions\n")
```

```{r umap-persons, fig.width=9, fig.height=7}
# Annotate with person-level usage stats (dominant category, sequence length)
person_stats <- df_tokens |>
  group_by(participant_code) |>
  summarise(
    seq_len       = n(),
    dominant_cat  = names(which.max(table(app_category))),
    pct_social    = mean(app_category == "social network"),
    pct_comm      = mean(app_category == "communication"),
    .groups = "drop"
  )

set.seed(42)
umap_out <- tryCatch(
  umap(person_emb, n_neighbors = min(15, max(2, nrow(person_emb) - 1))),
  error = function(e) {
    message("UMAP requires ≥ 3 persons — using PCA fallback: ", conditionMessage(e))
    NULL
  }
)

if (!is.null(umap_out)) {
  coords <- umap_out$layout
} else {
  pca_out <- prcomp(person_emb, scale. = FALSE)
  coords  <- pca_out$x[, 1:min(2, ncol(pca_out$x)), drop = FALSE]
  if (ncol(coords) < 2) coords <- cbind(coords, 0)
}

umap_df <- tibble(
  participant_code = rownames(person_emb),
  umap1 = coords[, 1],
  umap2 = coords[, 2]
) |>
  left_join(person_stats, by = "participant_code")

cat_pal <- c(
  "social network" = "#F28E2B", "communication" = "#4E79A7",
  "tools" = "#59A14F", "entertainment" = "#B07AA1", "other" = "#BAB0AC"
)

umap_df |>
  ggplot(aes(umap1, umap2, colour = dominant_cat, size = seq_len)) +
  geom_point(alpha = 0.75) +
  scale_colour_manual(values = cat_pal, name = "Dominant\ncategory") +
  scale_size(range = c(1.5, 6), name = "Sequence\nlength") +
  labs(title = "Person2Vec embedding space (UMAP)",
       subtitle = "Each point is one participant; proximity = similar usage patterns",
       x = "UMAP 1", y = "UMAP 2") +
  theme_minimal()
```

### Cluster persons in embedding space

```{r person-clusters, fig.width=9, fig.height=7}
# DECISION POINT: choose k via silhouette (same approach as Method 1)
# Cap at n_persons - 1; skip clustering when fewer than 3 persons are present.
n_persons_emb <- nrow(person_emb)
k_range       <- if (n_persons_emb >= 3) 2:min(6, n_persons_emb - 1) else integer(0)

if (length(k_range) > 0) {
  sil_widths <- sapply(k_range, function(k) {
    cl  <- kmeans(person_emb, centers = k, nstart = 25, iter.max = 100)
    sil <- silhouette(cl$cluster, dist(person_emb))
    mean(sil[, 3])
  })

  tibble(k = k_range, avg_silhouette = sil_widths) |>
    ggplot(aes(k, avg_silhouette)) +
    geom_line() + geom_point(size = 2) +
    labs(title = "Person cluster quality by k (k-means on Doc2Vec embeddings)",
         x = "k", y = "Average silhouette width") +
    theme_minimal()
} else {
  sil_widths <- numeric(0)
  cat(sprintf("Too few persons (%d) for clustering — needs full dataset.\n", n_persons_emb))
}
```

```{r cluster-umap, fig.width=9, fig.height=7}
if (length(k_range) > 0) {
  k_chosen <- k_range[which.max(sil_widths)]
  cat("Chosen k:", k_chosen, "\n")

  set.seed(42)
  km <- kmeans(person_emb, centers = k_chosen, nstart = 25)

  umap_df |>
    mutate(cluster = factor(km$cluster)) |>
    ggplot(aes(umap1, umap2, colour = cluster, size = seq_len)) +
    geom_point(alpha = 0.8) +
    scale_colour_brewer(palette = "Set2", name = "Cluster") +
    scale_size(range = c(1.5, 6), name = "Sequence\nlength") +
    labs(title = paste0("Person clusters in Doc2Vec space (k = ", k_chosen, ")"),
         x = "UMAP 1", y = "UMAP 2") +
    theme_minimal()
} else {
  km <- NULL
  cat("Skipping cluster UMAP — not enough persons.\n")
}
```

```{r cluster-profiles}
# What characterises each cluster?
if (!is.null(km)) {
tibble(
  participant_code = rownames(person_emb),
  cluster = factor(km$cluster)
) |>
  left_join(person_stats, by = "participant_code") |>
  group_by(cluster) |>
  summarise(
    n          = n(),
    med_seq_len = median(seq_len),
    pct_social  = round(mean(pct_social) * 100, 1),
    pct_comm    = round(mean(pct_comm) * 100, 1),
    dominant    = names(which.max(table(dominant_cat))),
    .groups = "drop"
  ) |>
  knitr::kable(caption = "Cluster profiles")
} else {
  cat("Skipping cluster profiles — not enough persons.\n")
}
```

---

## User trajectory visualisation

Plot a single person's sequence as a trajectory through the embedding
space — each point is one app use, arrows show the temporal order.
Mirrors the quiver-plot visualisation in the reference.

```{r trajectory, fig.width=9, fig.height=7}
# Pick one person for the trajectory (choose a person with a long sequence)
trajectory_person <- person_stats |>
  slice_max(seq_len, n = 1) |>
  pull(participant_code)

# Get their token sequence and look up embeddings
traj_tokens <- df_tokens |>
  filter(participant_code == trajectory_person) |>
  arrange(start_time) |>
  pull(token_rich)

# Only keep tokens that are in the Word2Vec vocabulary
in_vocab <- traj_tokens %in% rownames(emb_mat)
traj_emb <- emb_mat[traj_tokens[in_vocab], , drop = FALSE]

if (nrow(traj_emb) >= 4) {
  # Project to 2D with PCA fitted on all embeddings
  pca_all  <- prcomp(emb_mat, scale. = TRUE)
  traj_2d  <- predict(pca_all, traj_emb)[, 1:2]
  all_2d   <- pca_all$x[, 1:2]

  # Background: all token embeddings
  bg_df <- as_tibble(all_2d, rownames = "token") |>
    mutate(
      app_name = map_chr(str_split(token, "_(?=morning|afternoon|evening|night)"), 1),
      app_category = app_cat_lookup[app_name]
    )

  # Trajectory data frame: one row per step
  traj_df <- as_tibble(traj_2d) |>
    rename(x = PC1, y = PC2) |>
    mutate(
      t        = row_number(),
      token    = traj_tokens[in_vocab],
      app_name = map_chr(str_split(token, "_(?=morning|afternoon|evening|night)"), 1)
    )

  ggplot() +
    # Background cloud of all token embeddings
    geom_point(data = bg_df, aes(PC1, PC2, colour = app_category),
               alpha = 0.2, size = 2) +
    # Trajectory arrows
    geom_segment(
      data = traj_df |> mutate(xend = lead(x), yend = lead(y)) |> filter(!is.na(xend)),
      aes(x = x, y = y, xend = xend, yend = yend, colour = factor(t %% 8)),
      arrow = arrow(length = unit(2.5, "mm"), type = "closed"),
      linewidth = 0.7, show.legend = FALSE
    ) +
    # Points
    geom_point(data = traj_df, aes(x, y), colour = "black", size = 1.5) +
    scale_colour_manual(
      values = c("social network" = "#F28E2B", "communication" = "#4E79A7",
                 "tools" = "#59A14F", "entertainment" = "#B07AA1",
                 "other" = "#BAB0AC",
                 # trajectory step colours (recycled, hidden from legend)
                 setNames(RColorBrewer::brewer.pal(8, "PuRd"), as.character(0:7))),
      name = "Category"
    ) +
    labs(
      title = paste0("Trajectory of participant ", trajectory_person,
                     " through activity embedding space"),
      subtitle = "Arrows show temporal order of app uses",
      x = "PC1", y = "PC2"
    ) +
    theme_minimal() +
    guides(colour = guide_legend(override.aes = list(size = 4, alpha = 1)))
} else {
  cat("Not enough vocabulary overlap for trajectory plot.\n")
}
```

---

## Summary of key outputs

| Output | What it shows |
|--------|--------------|
| Vocabulary & token counts | How rich the token space is with/without time enrichment |
| Nearest neighbours table | Which apps appear in similar sequential contexts |
| t-SNE embedding plot | Activity embedding space — categories should cluster |
| UMAP person plot | Person-level behavioural space, coloured by dominant category |
| Silhouette plot | How many person clusters are supported |
| Cluster UMAP | Person clusters in embedding space |
| Cluster profiles | Usage characteristics of each behavioural cluster |
| Trajectory plot | One person's path through the embedding space over time |
